### 2019-02-11 extended-abstract.tex
added bibliography, extended abstract

%
% The first command in your LaTeX source must be the \documentclass command.
\documentclass[sigconf]{acmart}

%
% defining the \BibTeX command - from Oren Patashnik's original BibTeX documentation.
\def\BibTeX{{\rm B\kern-.05em{\sc i\kern-.025em b}\kern-.08emT\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}

% Rights management information. 
% This information is sent to you when you complete the rights form.
% These commands have SAMPLE values in them; it is your responsibility as an author to replace
% the commands and values with those provided to you when you complete the rights form.
%
% These commands are for a PROCEEDINGS abstract or paper.
\copyrightyear{2019}
\acmYear{2019}
\setcopyright{acmlicensed}
\acmConference[PEARC19]{PEARC19: Practice and Experience in Advanced Research Computing}{July 28 -- August 01, 2019}{Chicago, IL}
%\acmBooktitle{Woodstock '18: ACM Symposium on Neural Gaze Detection, June 03--05, 2018, Woodstock, NY}
%\acmPrice{15.00}
%\acmDOI{10.1145/1122445.1122456}
%\acmISBN{978-1-4503-9999-9/18/06}

%
% These commands are for a JOURNAL article.
%\setcopyright{acmcopyright}
%\acmJournal{TOG}
%\acmYear{2018}\acmVolume{37}\acmNumber{4}\acmArticle{111}\acmMonth{8}
%\acmDOI{10.1145/1122445.1122456}

%
% Submission ID. 
% Use this when submitting an article to a sponsored event. You'll receive a unique submission ID from the organizers
% of the event, and this ID should be used as the parameter to this command.
%\acmSubmissionID{123-A56-BU3}

%
% The majority of ACM publications use numbered citations and references. If you are preparing content for an event
% sponsored by ACM SIGGRAPH, you must use the "author year" style of citations and references. Uncommenting
% the next command will enable that style.
%\citestyle{acmauthoryear}

%
% end of the preamble, start of the body of the document source.
\begin{document}

%
% The "title" command has an optional parameter, allowing the author to define a "short title" to be used in page headers.
\title{Providing A Unified User Environment for Canada’s National Advanced Computing Centers}

%
% The "author" command and its associated commands are used to define the authors and their affiliations.
% Of note is the shared affiliation of the first two authors, and the "authornote" and "authornotemark" commands
% used to denote shared contribution to the research.
\author{Maxime Boissonneault}
\email{maxime.boissonneault@calculquebec.ca}
\affiliation{%
  \institution{Université Laval, Calcul Québec, Compute Canada}
  \streetaddress{2325 Rue de l'Université}
  \city{Québec}
  \state{Québec}
  \country{Canada}
  \postcode{G1V 0A6}
}
\author{Bart Oldeman}
\email{bart.oldeman@calculquebec.ca}
\affiliation{%
  \institution{McGill University, Calcul Québec, Compute Canada}
  \streetaddress{845 Rue Sherbrooke Ouest}
  \city{Montréal}
  \state{Québec}
  \country{Canada}
  \postcode{H3A 0G4}
}

\author{Ryan P. Taylor}
\email{rptaylor@uvic.ca}
\affiliation{%
  \institution{University of Victoria, WestGrid, Compute Canada}
  \streetaddress{3800 Finnerty Rd}
  \city{Victoria}
  \state{British Columbia}
  \country{Canada}
  \postcode{V8P 5C2}}

%
% By default, the full list of authors will be used in the page headers. Often, this list is too long, and will overlap
% other information printed in the page headers. This command allows the author to define a more concise list
% of authors' names for this purpose.
\renewcommand{\shortauthors}{Boissonneault, Oldeman and Taylor}

%
% The abstract is a short summary of the work to be presented in the article.
\begin{abstract}
Exploiting an advanced computing platform consisting of several clusters distributed across the second-largest country in the world is challenging. Each cluster may run a different operating system, use a different generation of CPU, GPU, or network fabric, or be managed by a different team of system administrators. Presenting a unified software environment can tremendously facilitate the task of supporting researchers, but is difficult to implement. This is nevertheless what Compute Canada set out to do in 2017, in the midst of deploying a new generation of large clusters.

In order to achieve this goal, we had to find software solutions to solve these challenges. Distribution, portability and performance were three important technical criteria for us. We also had to consider the practicality of each approach for our users, and reproducibility of software installations performed by staff located at various sites across Canada.

In this paper, we present the solution that we created, which has allowed Compute Canada to serve the needs of over 10,000 researchers across the country. It is used on over 20 different clusters with heterogeneous configurations, from CPU architectures as old as Nehalem or Opteron all the way up to Skylake, with and without GPUs, with InfiniBand, Ethernet or OmniPath as the network fabric, and with Slurm or Torque/Moab as the scheduler. It presents a unified software environment to the users, providing over 600 different scientific applications that are available in over 4,000 different combinations of version, compiler and CPU architecture.
\end{abstract}

%
% The code below is generated by the tool at http://dl.acm.org/ccs.cfm.
% Please copy and paste the code instead of the example below.
%

%\ccsdesc[500]{Computer systems organization~Embedded systems}
%\ccsdesc[300]{Computer systems organization~Redundancy}
%\ccsdesc{Computer systems organization~Robotics}
%\ccsdesc[100]{Networks~Network reliability}

%
% Keywords. The author(s) should pick words that accurately describe the work being
% presented. Separate the keywords with commas.
\keywords{cvmfs, easybuild, nix, software deployment}

%
% A "teaser" image appears between the author and affiliation information and the body 
% of the document, and typically spans the page. 
%\begin{teaserfigure}
%  \includegraphics[width=\textwidth]{sampleteaser}
%  \caption{Seattle Mariners at Spring Training, 2010.}
%  \Description{Enjoying the baseball game from the third-base seats. Ichiro Suzuki preparing to bat.}
%  \label{fig:teaser}
%\end{teaserfigure}

%
% This command processes the author and affiliation and title information and builds
% the first part of the formatted document.
\maketitle

\section{Introduction}
In 2017, Compute Canada set out to refresh its advanced computing infrastructure by deploying a few large clusters to replace nearly 30 smaller legacy clusters. This was a good opportunity to rethink our approach to software installation. Up to this point, each of the legacy clusters required individual manual software installation. This was laborious because each software package had to be installed tens of times, and it created a poor experience for users when moving between clusters, due to slight inconsistencies in software installations. When we set out to build a common software environment to be used on all new clusters, we first established four guiding principles for our design choices.

The first principle is that software packages should be available on every cluster (unless not permitted because of license or hardware reasons). This means that we need a scalable resilient multi-site distribution mechanism. There are a few solutions that could be used, and we will discuss them in Section 1. The solution we opted for is Cern Virtual Machine File System, or CVMFS\cite{CVMFS}.

The second principle is that the software environment should be decoupled from the underlying operating system. When we first designed the solution, new clusters were being deployed with the CentOS 7 operating system. However, there was interest to support some clusters running CentOS 6. Moreover, not every CentOS 7 system has the same set of system packages installed. Finally, Compute Canada also operates several OpenStack clouds, on which users can launch virtual machines running any Linux distribution. In order to use the unified software environment on virtual machines and multiple clusters, we needed a compatibility layer. The solution we initially deployed is Nix \cite{Nix}, although we are experimenting with Gentoo Prefix \cite{Gentoo} as a replacement because of issues we encountered with Nix. We will discuss these options in Section 2.

The third principle is that installation procedures should be tracked and reproducible. This means that automation is needed. The solution identified for the compatibility layer already addresses this issue. However, these solutions mostly target basic Linux packages. For scientific packages, we also wanted to ensure that the software packages were optimized for specific hardware architectures. Two tools with a large set of existing recipes were known to us at the time: Spack\cite{Spack} and EasyBuild \cite{EasyBuild2012,EasyBuild2014,EasyBuild2016}.

The fourth principle is that the interface presented to users should be as familiar as possible, and remain manageable with a very large number of software packages installed. All of our clusters were using a system of environment modules \cite{Modules1991,Modules1996}. However, the traditional systems of modules become unwieldy to handle at the scale of thousands of installed modules. Using a hierarchical module tree, which is possible in a newer implementation of environment modules called Lmod \cite{Lmod}, helps address this problem. We will discuss how our implementation meets the third and fourth principles in Section 3.

\section{Distribution mechanism}
Installing software packages on a single cluster can be done through a shared file system, such as NFS, Lustre, or GPFS. Anyone who has been working in the field of high performance computing for a number of years knows that parallel file systems are often a bottleneck, especially for the IO pattern caused by the action of launching applications. In addition to that problem, this solution is also a single point of failure, since any failure of that shared file system will block the vast majority of jobs. Another solution for a single cluster is to install binary packages on each node, using a mechanism such as RPM. This solution can however lead to diverging configurations on different nodes, and require an additional layer to keep the configurations synchronized.

Adding multiple clusters to the problem poses a second synchronization problem. To keep clusters synchronized, there are two strategies. The first one is to push changes from a central repository to the clusters. This can be done with a tool such as rsync. This however assumes that the target clusters are online whenever a change needs to be propagated, which is whenever a software package is installed or updated. The second solution is for the clusters to pull changes from the central repository on a schedule. This too could be implemented with rsync, but for a large software stack, it can take a few hours. It also has no way to know whether a change is needed for the users on that cluster.

CERN Virtual Machine File System (CVMFS) \cite{CVMFS} addresses the above issues. It is a file system initially designed to distribute software packages to an array of clusters geographically distribute around the world. It is a resilient file system with no single point of failure, and it delivers files to the clients in a pull fashion : each client pulls the files it needs when it needs them. Multiple node-wise, cluster-wise and global cache layers ensure that pulling files is efficient.

\subsection{CVMFS structure}
CVMFS is designed similarly to the Network Time Protocol stratum model. There is one source of truth, which is called the stratum 0. This is the injection point for new software packages that are to be deployed on the cluster. This server is never accessed directly by the end users or the clusters. This means that if it has a failure, the only consequence is that one can not inject new files until the server is back.

The repository hosted by the stratum 0 is versioned. This means that it is possible to go back to a previous version, should the need arise. Each file and folder hosted in that repository is hashed. This allows for deduplication of contents and significant space saving, since multiple copies of the same file will only use as much space as one file. Files are also compressed for further space saving. The hash mechanism also allows CVMFS clients to check quickly for the validity of their content, and to invalidate their cache when it is outdated.

The complete content of the stratum 0 is replicated onto a set of multiple strata 1 servers. They provide redundancy and ensure that there is no single point of failure. These are the clusters’ access point to the software stack. Files are accessed through the HTTP protocol, which means that clusters can also configure a set of squid servers to ensure a cluster-specific local cache, which protects against a failure of the WAN connection. Finally, if the compute nodes have a local disk, they host a partial cache, to keep most recently used files local. Alternatively, a cache can be setup on a shared file system if compute nodes do not have local disks. This can however become a bottleneck and is a single point of failure.

\section{Compatibility layer}
In order to detach the software stack from the underlying operating system, we need a layer that allows to build almost everything, down to the C library. The only exclusions that we do not include are the Linux kernel, kernel modules, drivers, and anything that is to be run as a privileged user.

\subsection{Nix}
Nix \cite{Nix} is both a package manager and an operating system (NixOS). Its core principle is reproducibility. Therefore, the recipes attempt very hard and mostly succeed to not depend on the underlying operating system at all. Every software built with Nix is installed inside a folder named by the hash of its recipe and that of its dependencies. This ensures very strong reproducibility, but it also means that updating a package in place is virtually impossible. While reproducibility is a major consideration in advanced computing, we sometimes need to fix bugs or security issues, which is made very hard by the design choices in Nix. At the time of writing this paper, we are still using Nix as a compatibility layer, but we are investigating a switch to Gentoo Prefix described next.

\subsection{Gentoo Prefix}

\section{Scientific applications}

\subsection{Compiled software packages}
The majority of our software stack is made up of scientific software that are performance critical. This implies that they cannot be installed in the compatibility layer described above, because different clusters support different sets of CPU instructions. To install these packages, we use a package manager called EasyBuild \cite{EasyBuild2012,EasyBuild2014,EasyBuild2016}.

EasyBuild initially originated from University of Ghent, but has now over 200 contributors from all around the world. The framework is based on scripted recipes to install packages, usually from source code. It is designed to produce optimized binaries, and allows to easily recompile different versions of a given software using different toolchains, where a toolchain is composed of a compiler, an implementation of MPI, mathematical libraries and CUDA.

EasyBuild also automatically creates module files. Out of the box, EasyBuild supports flat naming schemes as well as hierarchical naming schemes. We have customized our own module naming scheme based on the hierarchical naming scheme. Hierarchical module trees offer the advantage of hiding modules that are incompatible with toolchains that are currently loaded. This can however throw off some users, since they don’t automatically see all available modules. Lmod \cite{Lmod} alleviates this issue by providing a command to crawl the full module tree and instructs users on how to load specific modules that might be available only with specific toolchains.

\subsection{Handling multiple CPU architectures}
In order to handle multiple CPU architectures, we have configured EasyBuild to install software packages in a different directory and module tree based on the architecture requested. There is also a wide selection of software packages that are installed in an architecture-independent fashion. This is the case, for example, for binary packages, packages that are not performance critical, or packages that consists of source files only.

When a software package is built for multiple architectures, the same recipe is used for all of them. While each cluster will typically show only a single architecture, users are allowed to select a different architecture if they wish to run exactly the same binary as on another cluster.

\subsection{Packages for scripting languages}
Whenever a software package supports the installation of subpackages, as is the case with the Python, R, or Perl scripting languages, we leverage the existing package manager of that language, rather than installing subpackages as modules. In the case of Python, we are building Python wheels [11] against our software stack. Wheels are binary packages that can be installed in a matter of seconds, similar to CentOS RPM packages. Since they are built by our team, and linked against our optimize software stack, we ensure that they are both easy to installed and offer good performance.

\subsection{Licensed software packages}
Compute Canada also has agreements with a number of software vendors or developers of software packages that require a license. In these cases, we install the package using the same tools, but it is installed on a separate CVMFS repository which is only accessible to Compute Canada clusters.

\section{Other benefits}
Because of the portability of our software stack and the distribution mechanism, it can be used on any cluster, or any virtual machine on a cloud. For example, for training events, we have been using it to provision small sized replicas of Compute Canada clusters in short-lived virtual clusters deployed on our OpenStack platform. We have also used this portable software environment to expose a cluster environment directly through a browser, via JupyterHub.

\section{Acknowledgments}

We thank the following organizations and individuals for sharing their time and expertise:
\begin{itemize}
	\item All of the Compute Canada team members who make it possible to manage such a large software stack
	\item Robert McLay, from Texas Advanced Computing Center, for always providing quick and efficient support and debugging for Lmod
	\item Kenneth Hoste from Ghent University, and Åke Sandgren from Umeå University, for many discussions regarding EasyBuild
	\item Kuang Chung Chen and Bryan Caron, from McGill University, for initiating the groundwork that lead to this project
	\item Dave Dykstra, Jakob Blomer, and the CernVM project team for support and development of CVMFS
\end{itemize}

%
% The next two lines define the bibliography style to be used, and the bibliography file.
\bibliographystyle{ACM-Reference-Format}
\bibliography{biblio}

\end{document}

### 2019-02-12 extended-abstract.tex
added reference for python wheels
changed year in the abstract

@@
 Presenting a unified software environment can tremendously facilitate the task of supporting researchers, but is difficult to implement.
-This is nevertheless what Compute Canada set out to do in 2017, in the midst of deploying a new generation of large clusters.
+This is nevertheless what Compute Canada set out to do in 2016, in the midst of deploying a new generation of large clusters.
 
@@
 Whenever a software package supports the installation of subpackages, as is the case with the Python, R, or Perl scripting languages, we leverage the existing package manager of that language, rather than installing subpackages as modules.
-In the case of Python, we are building Python wheels [11] against our software stack.
+In the case of Python, we are building Python wheels \cite{Wheels} against our software stack.
 Wheels are binary packages that can be installed in a matter of seconds, similar to CentOS RPM packages.

### 2019-02-12 extended-abstract.tex
wrapping text at 120 columns

@@
 \begin{abstract}
-Exploiting an advanced computing platform consisting of several clusters distributed across the second-largest country in the world is challenging.
-Each cluster may run a different operating system, use a different generation of CPU, GPU, or network fabric, or be managed by a different team of system administrators.
-Presenting a unified software environment can tremendously facilitate the task of supporting researchers, but is difficult to implement.
-This is nevertheless what Compute Canada set out to do in 2016, in the midst of deploying a new generation of large clusters.
+Exploiting an advanced computing platform consisting of several clusters distributed across the second-largest country
+in the world is challenging. Each cluster may run a different operating system, use a different generation of CPU, GPU,
+or network fabric, or be managed by a different team of system administrators. Presenting a unified software
+environment can tremendously facilitate the task of supporting researchers, but is difficult to implement. This is
+nevertheless what Compute Canada set out to do in 2016, in the midst of deploying a new generation of large clusters.
 
-In order to achieve this goal, we had to find software solutions to solve these challenges.
-Distribution, portability and performance were three important technical criteria for us.
-We also had to consider the practicality of each approach for our users, and reproducibility of software installations performed by staff located at various sites across Canada.
+In order to achieve this goal, we had to find software solutions to solve these challenges. Distribution, portability
+and performance were three important technical criteria for us. We also had to consider the practicality of each
+approach for our users, and reproducibility of software installations performed by staff located at various sites
+across Canada.
 
-In this paper, we present the solution that we created, which has allowed Compute Canada to serve the needs of over 10,000 researchers across the country.
-It is used on over 20 different clusters with heterogeneous configurations, from CPU architectures as old as Nehalem or Opteron all the way up to Skylake, with and without GPUs, with InfiniBand, Ethernet or OmniPath as the network fabric, and with Slurm or Torque/Moab as the scheduler.
-It presents a unified software environment to the users, providing over 600 different scientific applications that are available in over 4,000 different combinations of version, compiler and CPU architecture.
+In this paper, we present the solution that we created, which has allowed Compute Canada to serve the needs of over
+10,000 researchers across the country. It is used on over 20 different clusters with heterogeneous configurations, from
+CPU architectures as old as Nehalem or Opteron all the way up to Skylake, with and without GPUs, with InfiniBand,
+Ethernet or OmniPath as the network fabric, and with Slurm or Torque/Moab as the scheduler. It presents a unified
+software environment to the users, providing over 600 different scientific applications that are available in over
+4,000 different combinations of version, compiler and CPU architecture.
 \end{abstract}
@@
 \section{Introduction}
-In 2017, Compute Canada set out to refresh its advanced computing infrastructure by deploying a few large clusters to replace nearly 30 smaller legacy clusters.
-This was a good opportunity to rethink our approach to software installation.
-Up to this point, each of the legacy clusters required individual manual software installation.
-This was laborious because each software package had to be installed tens of times, and it created a poor experience for users when moving between clusters, due to slight inconsistencies in software installations.
-When we set out to build a common software environment to be used on all new clusters, we first established four guiding principles for our design choices.
+In 2017, Compute Canada set out to refresh its advanced computing infrastructure by deploying a few large clusters to
+replace nearly 30 smaller legacy clusters. This was a good opportunity to rethink our approach to software
+installation. Up to this point, each of the legacy clusters required individual manual software installation. This was
+laborious because each software package had to be installed tens of times, and it created a poor experience for users
+when moving between clusters, due to slight inconsistencies in software installations. When we set out to build a
+common software environment to be used on all new clusters, we first established four guiding principles for our design
+choices.
 
-The first principle is that software packages should be available on every cluster (unless not permitted because of license or hardware reasons).
-This means that we need a scalable resilient multi-site distribution mechanism.
-There are a few solutions that could be used, and we will discuss them in Section 1.
-The solution we opted for is Cern Virtual Machine File System, or CVMFS\cite{CVMFS}.
+The first principle is that software packages should be available on every cluster (unless not permitted because of
+license or hardware reasons). This means that we need a scalable resilient multi-site distribution mechanism. There are
+a few solutions that could be used, and we will discuss them in Section 1. The solution we opted for is Cern Virtual
+Machine File System, or CVMFS\cite{CVMFS}.
 
-The second principle is that the software environment should be decoupled from the underlying operating system.
-When we first designed the solution, new clusters were being deployed with the CentOS 7 operating system.
-However, there was interest to support some clusters running CentOS 6.
-Moreover, not every CentOS 7 system has the same set of system packages installed.
-Finally, Compute Canada also operates several OpenStack clouds, on which users can launch virtual machines running any Linux distribution.
-In order to use the unified software environment on virtual machines and multiple clusters, we needed a compatibility layer.
-The solution we initially deployed is Nix \cite{Nix}, although we are experimenting with Gentoo Prefix \cite{Gentoo} as a replacement because of issues we encountered with Nix.
-We will discuss these options in Section 2.
+The second principle is that the software environment should be decoupled from the underlying operating system. When we
+first designed the solution, new clusters were being deployed with the CentOS 7 operating system. However, there was
+interest to support some clusters running CentOS 6. Moreover, not every CentOS 7 system has the same set of system
+packages installed. Finally, Compute Canada also operates several OpenStack clouds, on which users can launch virtual
+machines running any Linux distribution. In order to use the unified software environment on virtual machines and
+multiple clusters, we needed a compatibility layer. The solution we initially deployed is Nix \cite{Nix}, although we
+are experimenting with Gentoo Prefix \cite{Gentoo} as a replacement because of issues we encountered with Nix. We will
+discuss these options in Section 2.
 
-The third principle is that installation procedures should be tracked and reproducible.
-This means that automation is needed.
-The solution identified for the compatibility layer already addresses this issue.
-However, these solutions mostly target basic Linux packages.
-For scientific packages, we also wanted to ensure that the software packages were optimized for specific hardware architectures.
-Two tools with a large set of existing recipes were known to us at the time: Spack\cite{Spack} and EasyBuild \cite{EasyBuild2012,EasyBuild2014,EasyBuild2016}.
+The third principle is that installation procedures should be tracked and reproducible. This means that automation is
+needed. The solution identified for the compatibility layer already addresses this issue. However, these solutions
+mostly target basic Linux packages. For scientific packages, we also wanted to ensure that the software packages were
+optimized for specific hardware architectures. Two tools with a large set of existing recipes were known to us at the
+time: Spack\cite{Spack} and EasyBuild \cite{EasyBuild2012,EasyBuild2014,EasyBuild2016}.
 
-The fourth principle is that the interface presented to users should be as familiar as possible, and remain manageable with a very large number of software packages installed.
-All of our clusters were using a system of environment modules \cite{Modules1991,Modules1996}.
-However, the traditional systems of modules become unwieldy to handle at the scale of thousands of installed modules.
-Using a hierarchical module tree, which is possible in a newer implementation of environment modules called Lmod \cite{Lmod}, helps address this problem.
-We will discuss how our implementation meets the third and fourth principles in Section 3.
+The fourth principle is that the interface presented to users should be as familiar as possible, and remain manageable
+with a very large number of software packages installed. All of our clusters were using a system of environment modules
+\cite{Modules1991,Modules1996}. However, the traditional systems of modules become unwieldy to handle at the scale of
+thousands of installed modules. Using a hierarchical module tree, which is possible in a newer implementation of
+environment modules called Lmod \cite{Lmod}, helps address this problem. We will discuss how our implementation meets
+the third and fourth principles in Section 3.
 
 \section{Distribution mechanism}
-Installing software packages on a single cluster can be done through a shared file system, such as NFS, Lustre, or GPFS.
-Anyone who has been working in the field of high performance computing for a number of years knows that parallel file systems are often a bottleneck, especially for the IO pattern caused by the action of launching applications.
-In addition to that problem, this solution is also a single point of failure, since any failure of that shared file system will block the vast majority of jobs.
-Another solution for a single cluster is to install binary packages on each node, using a mechanism such as RPM.
-This solution can however lead to diverging configurations on different nodes, and require an additional layer to keep the configurations synchronized.
+Installing software packages on a single cluster can be done through a shared file system, such as NFS, Lustre, or
+GPFS. Anyone who has been working in the field of high performance computing for a number of years knows that parallel
+file systems are often a bottleneck, especially for the IO pattern caused by the action of launching applications. In
+addition to that problem, this solution is also a single point of failure, since any failure of that shared file system
+will block the vast majority of jobs. Another solution for a single cluster is to install binary packages on each node,
+using a mechanism such as RPM. This solution can however lead to diverging configurations on different nodes, and
+require an additional layer to keep the configurations synchronized.
 
-Adding multiple clusters to the problem poses a second synchronization problem.
-To keep clusters synchronized, there are two strategies.
-The first one is to push changes from a central repository to the clusters.
-This can be done with a tool such as rsync.
-This however assumes that the target clusters are online whenever a change needs to be propagated, which is whenever a software package is installed or updated.
-The second solution is for the clusters to pull changes from the central repository on a schedule.
-This too could be implemented with rsync, but for a large software stack, it can take a few hours.
-It also has no way to know whether a change is needed for the users on that cluster.
+Adding multiple clusters to the problem poses a second synchronization problem. To keep clusters synchronized, there
+are two strategies. The first one is to push changes from a central repository to the clusters. This can be done with a
+tool such as rsync. This however assumes that the target clusters are online whenever a change needs to be propagated,
+which is whenever a software package is installed or updated. The second solution is for the clusters to pull changes
+from the central repository on a schedule. This too could be implemented with rsync, but for a large software stack, it
+can take a few hours. It also has no way to know whether a change is needed for the users on that cluster.
 
-CERN Virtual Machine File System (CVMFS) \cite{CVMFS} addresses the above issues.
-It is a file system initially designed to distribute software packages to an array of clusters geographically distribute around the world.
-It is a resilient file system with no single point of failure, and it delivers files to the clients in a pull fashion : each client pulls the files it needs when it needs them.
-Multiple node-wise, cluster-wise and global cache layers ensure that pulling files is efficient.
+CERN Virtual Machine File System (CVMFS) \cite{CVMFS} addresses the above issues. It is a file system initially
+designed to distribute software packages to an array of clusters geographically distribute around the world. It is a
+resilient file system with no single point of failure, and it delivers files to the clients in a pull fashion : each
+client pulls the files it needs when it needs them. Multiple node-wise, cluster-wise and global cache layers ensure
+that pulling files is efficient.
 
 \subsection{CVMFS structure}
-CVMFS is designed similarly to the Network Time Protocol stratum model.
-There is one source of truth, which is called the stratum 0.
-This is the injection point for new software packages that are to be deployed on the cluster.
-This server is never accessed directly by the end users or the clusters.
-This means that if it has a failure, the only consequence is that one can not inject new files until the server is back.
+CVMFS is designed similarly to the Network Time Protocol stratum model. There is one source of truth, which is called
+the stratum 0. This is the injection point for new software packages that are to be deployed on the cluster. This
+server is never accessed directly by the end users or the clusters. This means that if it has a failure, the only
+consequence is that one can not inject new files until the server is back.
 
-The repository hosted by the stratum 0 is versioned.
-This means that it is possible to go back to a previous version, should the need arise.
-Each file and folder hosted in that repository is hashed.
-This allows for deduplication of contents and significant space saving, since multiple copies of the same file will only use as much space as one file.
-Files are also compressed for further space saving.
-The hash mechanism also allows CVMFS clients to check quickly for the validity of their content, and to invalidate their cache when it is outdated.
+The repository hosted by the stratum 0 is versioned. This means that it is possible to go back to a previous version,
+should the need arise. Each file and folder hosted in that repository is hashed. This allows for deduplication of
+contents and significant space saving, since multiple copies of the same file will only use as much space as one file.
+Files are also compressed for further space saving. The hash mechanism also allows CVMFS clients to check quickly for
+the validity of their content, and to invalidate their cache when it is outdated.
 
-The complete content of the stratum 0 is replicated onto a set of multiple strata 1 servers.
-They provide redundancy and ensure that there is no single point of failure.
-These are the clusters’ access point to the software stack.
-Files are accessed through the HTTP protocol, which means that clusters can also configure a set of squid servers to ensure a cluster-specific local cache, which protects against a failure of the WAN connection.
-Finally, if the compute nodes have a local disk, they host a partial cache, to keep most recently used files local.
-Alternatively, a cache can be setup on a shared file system if compute nodes do not have local disks.
-This can however become a bottleneck and is a single point of failure.
+The complete content of the stratum 0 is replicated onto a set of multiple strata 1 servers. They provide redundancy
+and ensure that there is no single point of failure. These are the clusters’ access point to the software stack. Files
+are accessed through the HTTP protocol, which means that clusters can also configure a set of squid servers to ensure a
+cluster-specific local cache, which protects against a failure of the WAN connection. Finally, if the compute nodes
+have a local disk, they host a partial cache, to keep most recently used files local. Alternatively, a cache can be
+setup on a shared file system if compute nodes do not have local disks. This can however become a bottleneck and is a
+single point of failure.
 
 \section{Compatibility layer}
-In order to detach the software stack from the underlying operating system, we need a layer that allows to build almost everything, down to the C library.
-The only exclusions that we do not include are the Linux kernel, kernel modules, drivers, and anything that is to be run as a privileged user.
+In order to detach the software stack from the underlying operating system, we need a layer that allows to build almost
+everything, down to the C library. The only exclusions that we do not include are the Linux kernel, kernel modules,
+drivers, and anything that is to be run as a privileged user.
 
 \subsection{Nix}
-Nix \cite{Nix} is both a package manager and an operating system (NixOS).
-Its core principle is reproducibility.
+Nix \cite{Nix} is both a package manager and an operating system (NixOS). Its core principle is reproducibility.
 Therefore, the recipes attempt very hard and mostly succeed to not depend on the underlying operating system at all.
-Every software built with Nix is installed inside a folder named by the hash of its recipe and that of its dependencies.
-This ensures very strong reproducibility, but it also means that updating a package in place is virtually impossible.
-While reproducibility is a major consideration in advanced computing, we sometimes need to fix bugs or security issues, which is made very hard by the design choices in Nix.
-At the time of writing this paper, we are still using Nix as a compatibility layer, but we are investigating a switch to Gentoo Prefix described next.
+Every software built with Nix is installed inside a folder named by the hash of its recipe and that of its
+dependencies. This ensures very strong reproducibility, but it also means that updating a package in place is virtually
+impossible. While reproducibility is a major consideration in advanced computing, we sometimes need to fix bugs or
+security issues, which is made very hard by the design choices in Nix. At the time of writing this paper, we are still
+using Nix as a compatibility layer, but we are investigating a switch to Gentoo Prefix described next.
 
@@
 \subsection{Compiled software packages}
-The majority of our software stack is made up of scientific software that are performance critical.
-This implies that they cannot be installed in the compatibility layer described above, because different clusters support different sets of CPU instructions.
-To install these packages, we use a package manager called EasyBuild \cite{EasyBuild2012,EasyBuild2014,EasyBuild2016}.
+The majority of our software stack is made up of scientific software that are performance critical. This implies that
+they cannot be installed in the compatibility layer described above, because different clusters support different sets
+of CPU instructions. To install these packages, we use a package manager called EasyBuild
+\cite{EasyBuild2012,EasyBuild2014,EasyBuild2016}.
 
 EasyBuild initially originated from University of Ghent, but has now over 200 contributors from all around the world.
-The framework is based on scripted recipes to install packages, usually from source code.
-It is designed to produce optimized binaries, and allows to easily recompile different versions of a given software using different toolchains, where a toolchain is composed of a compiler, an implementation of MPI, mathematical libraries and CUDA.
+The framework is based on scripted recipes to install packages, usually from source code. It is designed to produce
+optimized binaries, and allows to easily recompile different versions of a given software using different toolchains,
+where a toolchain is composed of a compiler, an implementation of MPI, mathematical libraries and CUDA.
 
-EasyBuild also automatically creates module files.
-Out of the box, EasyBuild supports flat naming schemes as well as hierarchical naming schemes.
-We have customized our own module naming scheme based on the hierarchical naming scheme.
-Hierarchical module trees offer the advantage of hiding modules that are incompatible with toolchains that are currently loaded.
-This can however throw off some users, since they don’t automatically see all available modules.
-Lmod \cite{Lmod} alleviates this issue by providing a command to crawl the full module tree and instructs users on how to load specific modules that might be available only with specific toolchains.
+EasyBuild also automatically creates module files. Out of the box, EasyBuild supports flat naming schemes as well as
+hierarchical naming schemes. We have customized our own module naming scheme based on the hierarchical naming scheme.
+Hierarchical module trees offer the advantage of hiding modules that are incompatible with toolchains that are
+currently loaded. This can however throw off some users, since they don’t automatically see all available modules. Lmod
+\cite{Lmod} alleviates this issue by providing a command to crawl the full module tree and instructs users on how to
+load specific modules that might be available only with specific toolchains.
 
 \subsection{Handling multiple CPU architectures}
-In order to handle multiple CPU architectures, we have configured EasyBuild to install software packages in a different directory and module tree based on the architecture requested.
-There is also a wide selection of software packages that are installed in an architecture-independent fashion.
-This is the case, for example, for binary packages, packages that are not performance critical, or packages that consists of source files only.
+In order to handle multiple CPU architectures, we have configured EasyBuild to install software packages in a different
+directory and module tree based on the architecture requested. There is also a wide selection of software packages that
+are installed in an architecture-independent fashion. This is the case, for example, for binary packages, packages that
+are not performance critical, or packages that consists of source files only.
 
-When a software package is built for multiple architectures, the same recipe is used for all of them.
-While each cluster will typically show only a single architecture, users are allowed to select a different architecture if they wish to run exactly the same binary as on another cluster.
+When a software package is built for multiple architectures, the same recipe is used for all of them. While each
+cluster will typically show only a single architecture, users are allowed to select a different architecture if they
+wish to run exactly the same binary as on another cluster.
 
 \subsection{Packages for scripting languages}
-Whenever a software package supports the installation of subpackages, as is the case with the Python, R, or Perl scripting languages, we leverage the existing package manager of that language, rather than installing subpackages as modules.
-In the case of Python, we are building Python wheels \cite{Wheels} against our software stack.
-Wheels are binary packages that can be installed in a matter of seconds, similar to CentOS RPM packages.
-Since they are built by our team, and linked against our optimize software stack, we ensure that they are both easy to installed and offer good performance.
+Whenever a software package supports the installation of subpackages, as is the case with the Python, R, or Perl
+scripting languages, we leverage the existing package manager of that language, rather than installing subpackages as
+modules. In the case of Python, we are building Python wheels \cite{Wheels} against our software stack. Wheels are
+binary packages that can be installed in a matter of seconds, similar to CentOS RPM packages. Since they are built by
+our team, and linked against our optimize software stack, we ensure that they are both easy to installed and offer good
+performance.
 
 \subsection{Licensed software packages}
-Compute Canada also has agreements with a number of software vendors or developers of software packages that require a license.
-In these cases, we install the package using the same tools, but it is installed on a separate CVMFS repository which is only accessible to Compute Canada clusters.
+Compute Canada also has agreements with a number of software vendors or developers of software packages that require a
+license. In these cases, we install the package using the same tools, but it is installed on a separate CVMFS
+repository which is only accessible to Compute Canada clusters.
 
 \section{Other benefits}
-Because of the portability of our software stack and the distribution mechanism, it can be used on any cluster, or any virtual machine on a cloud.
-For example, for training events, we have been using it to provision small sized replicas of Compute Canada clusters in short-lived virtual clusters deployed on our OpenStack platform.
-We have also used this portable software environment to expose a cluster environment directly through a browser, via JupyterHub.
+Because of the portability of our software stack and the distribution mechanism, it can be used on any cluster, or any
+virtual machine on a cloud. For example, for training events, we have been using it to provision small sized replicas
+of Compute Canada clusters in short-lived virtual clusters deployed on our OpenStack platform. We have also used this
+portable software environment to expose a cluster environment directly through a browser, via JupyterHub.
 
@@
 	\item All of the Compute Canada team members who make it possible to manage such a large software stack
-	\item Robert McLay, from Texas Advanced Computing Center, for always providing quick and efficient support and debugging for Lmod
-	\item Kenneth Hoste from Ghent University, and Åke Sandgren from Umeå University, for many discussions regarding EasyBuild
-	\item Kuang Chung Chen and Bryan Caron, from McGill University, for initiating the groundwork that lead to this project
+	\item Robert McLay, from Texas Advanced Computing Center, for always providing quick and efficient support and
+debugging for Lmod
+	\item Kenneth Hoste from Ghent University, and Åke Sandgren from Umeå University, for many discussions
+regarding EasyBuild
+	\item Kuang Chung Chen and Bryan Caron, from McGill University, for initiating the groundwork that lead to this
+project
 	\item Dave Dykstra, Jakob Blomer, and the CernVM project team for support and development of CVMFS

### 2019-02-12 extended-abstract.tex
added labels and updated references to sections

@@
 \documentclass[sigconf]{acmart}
-
+%\usepackage[french]{babel}
+%\selectlanguage{french}
+\usepackage[T1]{fontenc}
+\usepackage[utf8]{inputenc}
 %
@@
 % The "title" command has an optional parameter, allowing the author to define a "short title" to be used in page headers.
-\title{Providing A Unified User Environment for Canada’s National Advanced Computing Centers}
+\title{Providing A Unified User Environment for Canada National Advanced Computing Centers}
 
@@
 license or hardware reasons). This means that we need a scalable resilient multi-site distribution mechanism. There are
-a few solutions that could be used, and we will discuss them in Section 1. The solution we opted for is Cern Virtual
-Machine File System, or CVMFS\cite{CVMFS}.
+a few solutions that could be used, and we will discuss them in Section~\ref{sec:Distribution_mechanism}. The solution
+we opted for is Cern Virtual Machine File System, or CVMFS\cite{CVMFS}.
 
@@
 are experimenting with Gentoo Prefix \cite{Gentoo} as a replacement because of issues we encountered with Nix. We will
-discuss these options in Section 2.
+discuss these options in Section~\ref{sec:Compatibility_layer}.
 
@@
 environment modules called Lmod \cite{Lmod}, helps address this problem. We will discuss how our implementation meets
-the third and fourth principles in Section 3.
+the third and fourth principles in Section~\ref{sec:Scientific_applications}.
 
 \section{Distribution mechanism}
+\label{sec:Distribution_mechanism}
 Installing software packages on a single cluster can be done through a shared file system, such as NFS, Lustre, or
@@
 \subsection{CVMFS structure}
+\label{sub:CVMFS_structure}
 CVMFS is designed similarly to the Network Time Protocol stratum model. There is one source of truth, which is called
@@
 \section{Compatibility layer}
+\label{sec:Compatibility_layer}
 In order to detach the software stack from the underlying operating system, we need a layer that allows to build almost
@@
 \subsection{Nix}
+\label{sub:Nix}
 Nix \cite{Nix} is both a package manager and an operating system (NixOS). Its core principle is reproducibility.
@@
 \subsection{Gentoo Prefix}
+\label{sub:Gentoo_Prefix}
 
 \section{Scientific applications}
+\label{sec:Scientific_applications}
 
 \subsection{Compiled software packages}
+\label{sub:Compiled_software_packages}
 The majority of our software stack is made up of scientific software that are performance critical. This implies that
@@
 \subsection{Handling multiple CPU architectures}
+\label{sub:Handling_multiple_CPU_architectures}
 In order to handle multiple CPU architectures, we have configured EasyBuild to install software packages in a different
@@
 \subsection{Packages for scripting languages}
+\label{sub:Packages_for_scripting_languages}
 Whenever a software package supports the installation of subpackages, as is the case with the Python, R, or Perl
@@
 \subsection{Licensed software packages}
+\label{sub:Licensed_software_packages}
 Compute Canada also has agreements with a number of software vendors or developers of software packages that require a
@@
 \section{Other benefits}
+\label{sec:Other_benefits}
 Because of the portability of our software stack and the distribution mechanism, it can be used on any cluster, or any
@@
 \section{Acknowledgments}
-
+\label{sec:Acknowledgments}
 We thank the following organizations and individuals for sharing their time and expertise:

### 2019-02-12 extended-abstract.tex
Added section about Gentoo

@@
 \section{Introduction}
+\label{sec:Introduction}
 In 2017, Compute Canada set out to refresh its advanced computing infrastructure by deploying a few large clusters to
@@
 \label{sub:Gentoo_Prefix}
+Gentoo Prefix \cite{Gentoo} is an alternative solution to provide a compatibility layer.
+It allows to install in a non-default location most packages that make a typical Linux operating system.
+Contrary to Nix however, it installs packages in a structure that is much more similar to a standard Linux distribution.
+This is helpful when we need to update packages, and it minimizes problems when an application encounters a dependency that is not installed in a traditional structure.
+At the time of this writing, we are still experimenting with Gentoo Prefix, and it is not used in production.
 

### 2019-02-12 extended-abstract.tex
Disabled maketitle to put it in the format of an extended abstract, without authors, title, abstract, etc.
Rephrased the introduction and CVMFS section to make it shorter.

@@
 % the first part of the formatted document.
-\maketitle
+%\maketitle
 
@@
 \label{sec:Introduction}
-In 2017, Compute Canada set out to refresh its advanced computing infrastructure by deploying a few large clusters to
-replace nearly 30 smaller legacy clusters. This was a good opportunity to rethink our approach to software
-installation. Up to this point, each of the legacy clusters required individual manual software installation. This was
-laborious because each software package had to be installed tens of times, and it created a poor experience for users
-when moving between clusters, due to slight inconsistencies in software installations. When we set out to build a
-common software environment to be used on all new clusters, we first established four guiding principles for our design
-choices.
+In 2016, Compute Canada set out to refresh its advanced computing infrastructure by deploying a few large clusters to replace nearly 30 smaller legacy clusters.
+This was a good opportunity to rethink our approach to software installation.
+Up to this point, each of the legacy clusters required individual manual software installation.
+This was laborious because each software package had to be installed tens of times, and it created a poor experience for users when moving between clusters, due to inconsistencies in software installations.
+When we set out to build a common software environment, we first established four guiding principles for our design choices.
 
-The first principle is that software packages should be available on every cluster (unless not permitted because of
-license or hardware reasons). This means that we need a scalable resilient multi-site distribution mechanism. There are
-a few solutions that could be used, and we will discuss them in Section~\ref{sec:Distribution_mechanism}. The solution
-we opted for is Cern Virtual Machine File System, or CVMFS\cite{CVMFS}.
+The first principle is that software packages should be available on every cluster (unless not permitted because of license or hardware reasons).
+This means that we need a scalable resilient multi-site distribution mechanism.
+There are a few solutions that could be used, and we will discuss them in Section~\ref{sec:Distribution_mechanism}.
+The solution we opted for is Cern Virtual Machine File System, or CVMFS\cite{CVMFS}.
 
-The second principle is that the software environment should be decoupled from the underlying operating system. When we
-first designed the solution, new clusters were being deployed with the CentOS 7 operating system. However, there was
-interest to support some clusters running CentOS 6. Moreover, not every CentOS 7 system has the same set of system
-packages installed. Finally, Compute Canada also operates several OpenStack clouds, on which users can launch virtual
-machines running any Linux distribution. In order to use the unified software environment on virtual machines and
-multiple clusters, we needed a compatibility layer. The solution we initially deployed is Nix \cite{Nix}, although we
-are experimenting with Gentoo Prefix \cite{Gentoo} as a replacement because of issues we encountered with Nix. We will
-discuss these options in Section~\ref{sec:Compatibility_layer}.
+The second principle is that the software environment should be decoupled from the underlying operating system.
+When we first designed the solution, new clusters were being deployed with the CentOS 7 operating system.
+However, they did not all have the same set of system packages installed.
+Moreover, we were interested in supporting legacy clusters that were running CentOS 6, and virtual machines launched by users on one of several Compute Canada cloud platforms.
+In order to use the unified software environment on virtual machines and multiple clusters, we needed a compatibility layer.
+The solution we deployed is Nix \cite{Nix}, although we are experimenting with Gentoo Prefix \cite{Gentoo} as a replacement.
+We will discuss these options in Section~\ref{sec:Compatibility_layer}.
 
-The third principle is that installation procedures should be tracked and reproducible. This means that automation is
-needed. The solution identified for the compatibility layer already addresses this issue. However, these solutions
-mostly target basic Linux packages. For scientific packages, we also wanted to ensure that the software packages were
-optimized for specific hardware architectures. Two tools with a large set of existing recipes were known to us at the
-time: Spack\cite{Spack} and EasyBuild \cite{EasyBuild2012,EasyBuild2014,EasyBuild2016}.
+The third principle is that installation procedures should be tracked and reproducible.
+This means that automation is needed.
+The solution identified for the compatibility layer already addresses this issue.
+However, these solutions mostly target basic Linux packages.
+For scientific packages, we also wanted to ensure that the software packages were optimized for specific hardware architectures.
+Two tools with a large set of existing recipes were known to us at the time: Spack\cite{Spack} and EasyBuild \cite{EasyBuild2012,EasyBuild2014,EasyBuild2016}.
+We will discuss why we chose EasyBuild in Section~\ref{sub:Compiled_software_packages}.
 
-The fourth principle is that the interface presented to users should be as familiar as possible, and remain manageable
-with a very large number of software packages installed. All of our clusters were using a system of environment modules
-\cite{Modules1991,Modules1996}. However, the traditional systems of modules become unwieldy to handle at the scale of
-thousands of installed modules. Using a hierarchical module tree, which is possible in a newer implementation of
-environment modules called Lmod \cite{Lmod}, helps address this problem. We will discuss how our implementation meets
-the third and fourth principles in Section~\ref{sec:Scientific_applications}.
+The fourth principle is that the interface presented to users should be as familiar as possible, and remain manageable with a very large number of software packages installed.
+All of our clusters were using a system of environment modules \cite{Modules1991,Modules1996}.
+However, the traditional systems of modules become unwieldy to handle at the scale of thousands of installed modules.
+Using a hierarchical module tree, which is possible in a newer implementation of environment modules called Lmod \cite{Lmod}, helps address this problem.
+We will discuss how our implementation meets the third and fourth principles in Section~\ref{sec:Scientific_applications}.
 
@@
 \label{sec:Distribution_mechanism}
-Installing software packages on a single cluster can be done through a shared file system, such as NFS, Lustre, or
-GPFS. Anyone who has been working in the field of high performance computing for a number of years knows that parallel
-file systems are often a bottleneck, especially for the IO pattern caused by the action of launching applications. In
-addition to that problem, this solution is also a single point of failure, since any failure of that shared file system
-will block the vast majority of jobs. Another solution for a single cluster is to install binary packages on each node,
-using a mechanism such as RPM. This solution can however lead to diverging configurations on different nodes, and
-require an additional layer to keep the configurations synchronized.
+Installing software packages on a single cluster can be done through a shared file system, such as NFS, Lustre, or GPFS.
+Anyone who has been working in the field of high performance computing for a number of years knows that parallel file systems are often a bottleneck, especially for the IO pattern caused by the action of launching applications.
+In addition to that problem, this solution is also a single point of failure, since any failure of that shared file system will block the vast majority of jobs.
+Another solution for a single cluster is to install binary packages on each node, using a mechanism such as RPM.
+This solution can however lead to diverging configurations on different nodes, and require an additional layer to keep the configurations synchronized.
 
-Adding multiple clusters to the problem poses a second synchronization problem. To keep clusters synchronized, there
-are two strategies. The first one is to push changes from a central repository to the clusters. This can be done with a
-tool such as rsync. This however assumes that the target clusters are online whenever a change needs to be propagated,
-which is whenever a software package is installed or updated. The second solution is for the clusters to pull changes
-from the central repository on a schedule. This too could be implemented with rsync, but for a large software stack, it
-can take a few hours. It also has no way to know whether a change is needed for the users on that cluster.
+Adding multiple clusters to the problem poses a second synchronization problem.
+To keep clusters synchronized, there are two strategies : pushing changes to the clusters, or pulling changes from a central location.
+Both strategies could be implemented with a tool such as {\it rsync}.
+However, synchronizing directories with over 40 million files and  directories takes a few hours, even with a tool as efficient as {\it rsync}.
 
-CERN Virtual Machine File System (CVMFS) \cite{CVMFS} addresses the above issues. It is a file system initially
-designed to distribute software packages to an array of clusters geographically distribute around the world. It is a
-resilient file system with no single point of failure, and it delivers files to the clients in a pull fashion : each
-client pulls the files it needs when it needs them. Multiple node-wise, cluster-wise and global cache layers ensure
-that pulling files is efficient.
+We instead chose to use CERN Virtual Machine File System (CVMFS) \cite{CVMFS}.
+CVMFS is a file system initially designed to distribute software packages to an array of clusters geographically distributed around the world.
+It is a resilient file system with no single point of failure, and it delivers files to the clients in a pull fashion: each client pulls file from it as needed.
+Multiple cache layers, at the level of each node and each cluster, keep network traffic to the central repository to a minimum.
 
@@
 \label{sub:CVMFS_structure}
-CVMFS is designed similarly to the Network Time Protocol stratum model. There is one source of truth, which is called
-the stratum 0. This is the injection point for new software packages that are to be deployed on the cluster. This
-server is never accessed directly by the end users or the clusters. This means that if it has a failure, the only
-consequence is that one can not inject new files until the server is back.
+CVMFS is designed similarly to the Network Time Protocol stratum model.
+There is one source of truth called the stratum 0, where any injection of new software is done.
+The stratum 0 is never accessed directly by users or even by the clusters.
+If it fails, the only immediate consequence is that injection of new files is not possible until it is back online.
 
-The repository hosted by the stratum 0 is versioned. This means that it is possible to go back to a previous version,
-should the need arise. Each file and folder hosted in that repository is hashed. This allows for deduplication of
-contents and significant space saving, since multiple copies of the same file will only use as much space as one file.
-Files are also compressed for further space saving. The hash mechanism also allows CVMFS clients to check quickly for
-the validity of their content, and to invalidate their cache when it is outdated.
+Snapshots of the stratum 0 are kept for a while, such that it is possible to go back to a previous version if needed.
+Files in the repository are compressed, and their hash is computed to allow space saving by avoiding duplication.
+Hashesalso allows CVMFS clients to quickly invalidate their local cache as needed.
 
-The complete content of the stratum 0 is replicated onto a set of multiple strata 1 servers. They provide redundancy
-and ensure that there is no single point of failure. These are the clusters’ access point to the software stack. Files
-are accessed through the HTTP protocol, which means that clusters can also configure a set of squid servers to ensure a
-cluster-specific local cache, which protects against a failure of the WAN connection. Finally, if the compute nodes
-have a local disk, they host a partial cache, to keep most recently used files local. Alternatively, a cache can be
-setup on a shared file system if compute nodes do not have local disks. This can however become a bottleneck and is a
-single point of failure.
+The stratum 0 is replicated to multiple strata 1 servers that are the access points of the clusters to the software environment.
+They provide redundancy and avoid single points of failure.
+Files are served to the clusters through HTTP and can be cached locally with multiple proxy servers, protecting against an outage of the wide area network.
+The closest layer of cache can be configured either to use a local disk on the compute nodes, or on a shared file system on in a diskless environment.
 

### 2019-02-13 extended-abstract.tex
Rephrasing of the rest of the extended abstract.

@@
 \label{sec:Compatibility_layer}
-In order to detach the software stack from the underlying operating system, we need a layer that allows to build almost
-everything, down to the C library. The only exclusions that we do not include are the Linux kernel, kernel modules,
-drivers, and anything that is to be run as a privileged user.
+Isolating the software environment from the operating system requires a compatibility layer containing almost everything that an OS would contain.
+From this layer, we exclude the Linux kernel and its modules, hardware drivers as well as privileged binaries.
 
@@
 \label{sub:Nix}
-Nix \cite{Nix} is both a package manager and an operating system (NixOS). Its core principle is reproducibility.
-Therefore, the recipes attempt very hard and mostly succeed to not depend on the underlying operating system at all.
-Every software built with Nix is installed inside a folder named by the hash of its recipe and that of its
-dependencies. This ensures very strong reproducibility, but it also means that updating a package in place is virtually
-impossible. While reproducibility is a major consideration in advanced computing, we sometimes need to fix bugs or
-security issues, which is made very hard by the design choices in Nix. At the time of writing this paper, we are still
-using Nix as a compatibility layer, but we are investigating a switch to Gentoo Prefix described next.
+Nix \cite{Nix} is both a package manager and an operating system (NixOS).
+Reproducibility being one the core principles in Nix, it is designed to depend as little as possible on the underlying operating system.
+Every software package installed through nix resides in a path calculated with the hash of the recipe to build it.
+This ensures very strong reproducibility, but it also makes updating a package in place impossible.
+While reproducibility is an important consideration in advanced research computing, bugs sometimes need to be fixed, which is difficult due to Nix' design.
+At the time of writing this paper, we are still using Nix as a compatibility layer, but we are investigating a switch to Gentoo Prefix described next.
 
@@
 \label{sub:Gentoo_Prefix}
-Gentoo Prefix \cite{Gentoo} is an alternative solution to provide a compatibility layer.
-It allows to install in a non-default location most packages that make a typical Linux operating system.
-Contrary to Nix however, it installs packages in a structure that is much more similar to a standard Linux distribution.
-This is helpful when we need to update packages, and it minimizes problems when an application encounters a dependency that is not installed in a traditional structure.
+Gentoo Prefix~\cite{Gentoo} is an alternative solution to provide a compatibility layer.
+It allows installation of most packages that make a typical Linux inside a non-default location.
+Contrary to Nix however, packages are installed in a standard directory structure.
+This is helpful when updates are required, and minimizes problems with applications that expect a standard directory structure.
 At the time of this writing, we are still experimenting with Gentoo Prefix, and it is not used in production.
@@
 \label{sub:Compiled_software_packages}
-The majority of our software stack is made up of scientific software that are performance critical. This implies that
-they cannot be installed in the compatibility layer described above, because different clusters support different sets
-of CPU instructions. To install these packages, we use a package manager called EasyBuild
-\cite{EasyBuild2012,EasyBuild2014,EasyBuild2016}.
+Scientific applications constitute the majority of our software stack.
+Since their performance is critical, they can not be installed in the compatibility layer, as they need to be optimized for specific CPU instructions.
+To install them, we use the EasyBuild~\cite{EasyBuild2012,EasyBuild2014,EasyBuild2016} package manager.
 
-EasyBuild initially originated from University of Ghent, but has now over 200 contributors from all around the world.
-The framework is based on scripted recipes to install packages, usually from source code. It is designed to produce
-optimized binaries, and allows to easily recompile different versions of a given software using different toolchains,
-where a toolchain is composed of a compiler, an implementation of MPI, mathematical libraries and CUDA.
+EasyBuild originated from University of Ghent, but has now over 200 contributors from all around the world.
+The framework is based on scripted recipes to install packages, usually from source code.
+It is designed to create optimized binaries, and eases recompilation of multiple versions of a given application using many combinations of compilers or libraries.
 
-EasyBuild also automatically creates module files. Out of the box, EasyBuild supports flat naming schemes as well as
-hierarchical naming schemes. We have customized our own module naming scheme based on the hierarchical naming scheme.
-Hierarchical module trees offer the advantage of hiding modules that are incompatible with toolchains that are
-currently loaded. This can however throw off some users, since they don’t automatically see all available modules. Lmod
-\cite{Lmod} alleviates this issue by providing a command to crawl the full module tree and instructs users on how to
-load specific modules that might be available only with specific toolchains.
+EasyBuild also automatically creates module files~\cite{Modules1991,Modules1996,Lmod}.
+The framework supports multiple standard and custom naming schemes.
+We use a version of a hierarchical naming scheme, in which modules are organized in a tree structure, hiding those that are incompatible currently loaded modules.
+This limits the list of modules which could otherwise be very long.
+This can however throw off some users, since they don’t automatically see all available modules.
+Lmod~\cite{Lmod} alleviates this issue by providing a command to crawl the full module tree and instructs users on how to load the module they want.
 
@@
 \label{sub:Handling_multiple_CPU_architectures}
-In order to handle multiple CPU architectures, we have configured EasyBuild to install software packages in a different
-directory and module tree based on the architecture requested. There is also a wide selection of software packages that
-are installed in an architecture-independent fashion. This is the case, for example, for binary packages, packages that
-are not performance critical, or packages that consists of source files only.
-
-When a software package is built for multiple architectures, the same recipe is used for all of them. While each
-cluster will typically show only a single architecture, users are allowed to select a different architecture if they
-wish to run exactly the same binary as on another cluster.
+Handling multiple CPU architectures is done by installing applications in architecture dependent paths and module trees.
+We also have a wide selection of software packages that are installed in an architecture-independent fashion.
+This is the case, for example, for binary packages, packages that are not performance critical, or packages that consists of source files only.
+The same recipe is typically used for installing the same package for multiple architectures.
+While each cluster will usually show only a single architecture, users are allowed to select a different architecture if they wish to run exactly the same binary as on another cluster.
 
@@
 \label{sub:Packages_for_scripting_languages}
-Whenever a software package supports the installation of subpackages, as is the case with the Python, R, or Perl
-scripting languages, we leverage the existing package manager of that language, rather than installing subpackages as
-modules. In the case of Python, we are building Python wheels \cite{Wheels} against our software stack. Wheels are
-binary packages that can be installed in a matter of seconds, similar to CentOS RPM packages. Since they are built by
-our team, and linked against our optimize software stack, we ensure that they are both easy to installed and offer good
-performance.
+For scripting languages that support installing packages, such as Python, R or Perl, we leverage the existing package manager of that language, rather than installing them as modules.
+In the case of Python, we are building Python wheels~\cite{Wheels}, linking against our software stack.
+Wheels are binary packages that can be installed in a matter of seconds, similar to CentOS RPM packages.
+Since they are built by our team, and linked against our optimize software stack, we ensure that they are both easy to installed and offer good performance.
 
-\subsection{Licensed software packages}
+\subsection{Licensed and restricted software packages}
 \label{sub:Licensed_software_packages}
-Compute Canada also has agreements with a number of software vendors or developers of software packages that require a
-license. In these cases, we install the package using the same tools, but it is installed on a separate CVMFS
-repository which is only accessible to Compute Canada clusters.
+Compute Canada also has agreements with a number of software vendors or developers of applications that require a license.
+In these cases, we install the package using the same tools, but it is deployed on a separate CVMFS repository which is only accessible to Compute Canada clusters.
 
@@
 \label{sec:Other_benefits}
-Because of the portability of our software stack and the distribution mechanism, it can be used on any cluster, or any
-virtual machine on a cloud. For example, for training events, we have been using it to provision small sized replicas
-of Compute Canada clusters in short-lived virtual clusters deployed on our OpenStack platform. We have also used this
-portable software environment to expose a cluster environment directly through a browser, via JupyterHub.
+Because of the portability of our software stack and the distribution mechanism, it can be used on any cluster, or any virtual machine on a cloud.
+For example, for training events, we have been using it to provision small sized replicas of Compute Canada clusters in short-lived virtual clusters deployed on our OpenStack platform.
+We have also used this portable software environment to expose a cluster environment directly through a browser, via JupyterHub.
+It can also help support infrastructures owned by various institutions.
 

### 2019-02-18 extended-abstract.tex
Updates from Ryan and Kenneth

@@
 \documentclass[sigconf]{acmart}
-%\usepackage[french]{babel}
-%\selectlanguage{french}
 \usepackage[T1]{fontenc}
@@
 % The "title" command has an optional parameter, allowing the author to define a "short title" to be used in page headers.
-\title{Providing A Unified User Environment for Canada National Advanced Computing Centers}
+\title{Providing A Unified Software Environment for Canada’s National Advanced Computing Centers}
 
@@
 \begin{abstract}
-Exploiting an advanced computing platform consisting of several clusters distributed across the second-largest country
-in the world is challenging. Each cluster may run a different operating system, use a different generation of CPU, GPU,
-or network fabric, or be managed by a different team of system administrators. Presenting a unified software
-environment can tremendously facilitate the task of supporting researchers, but is difficult to implement. This is
-nevertheless what Compute Canada set out to do in 2016, in the midst of deploying a new generation of large clusters.
+Exploiting an advanced computing platform consisting of several clusters distributed across the second-largest country in the world is challenging.
+Each cluster may run a different operating system, use a different generation of CPU, GPU, or network fabric, or be managed by a different team of system administrators.
+Presenting a unified software environment can tremendously facilitate the task of supporting researchers, but is challenging to implement.
+This is nevertheless what Compute Canada set out to do in 2016, in the midst of deploying a new generation of large clusters.
 
-In order to achieve this goal, we had to find software solutions to solve these challenges. Distribution, portability
-and performance were three important technical criteria for us. We also had to consider the practicality of each
-approach for our users, and reproducibility of software installations performed by staff located at various sites
-across Canada.
+We had to find software solutions to solve the challenges involved to achieve this goal.
+Distribution, portability and performance were three important technical criteria for us.
+We also had to consider the practicality of each approach for our users, and reproducibility of software installations performed by staff located at various sites across Canada.
 
-In this paper, we present the solution that we created, which has allowed Compute Canada to serve the needs of over
-10,000 researchers across the country. It is used on over 20 different clusters with heterogeneous configurations, from
-CPU architectures as old as Nehalem or Opteron all the way up to Skylake, with and without GPUs, with InfiniBand,
-Ethernet or OmniPath as the network fabric, and with Slurm or Torque/Moab as the scheduler. It presents a unified
-software environment to the users, providing over 600 different scientific applications that are available in over
-4,000 different combinations of version, compiler and CPU architecture.
+In this paper, we present the solution that we created, which has allowed Compute Canada to serve the needs of over 10,000 researchers across the country.
+It is used on over 20 different clusters with heterogeneous configurations, from processor architectures as old as Intel Nehalem or AMD Opteron all the way up to Intel Skylake, with and without GPUs, with InfiniBand, Ethernet or OmniPath as the network fabric, and with Slurm or Torque/Moab as the scheduler.
+It presents a unified software environment to the users, providing over 600 different scientific applications that are available in over 4,000 different combinations of version, compiler and CPU architecture.
 \end{abstract}
@@
 Up to this point, each of the legacy clusters required individual manual software installation.
-This was laborious because each software package had to be installed tens of times, and it created a poor experience for users when moving between clusters, due to inconsistencies in software installations.
+This was laborious because each software package had to be installed dozens of times, and it created a poor experience for users when moving between clusters, due to inconsistencies in software installations.
 When we set out to build a common software environment, we first established four guiding principles for our design choices.
 
-The first principle is that software packages should be available on every cluster (unless not permitted because of license or hardware reasons).
-This means that we need a scalable resilient multi-site distribution mechanism.
+The first principle is that software packages should be available on every cluster (unless not possible because of license or hardware reasons).
+This meant that we needed a scalable resilient multi-site distribution mechanism.
 There are a few solutions that could be used, and we will discuss them in Section~\ref{sec:Distribution_mechanism}.
-The solution we opted for is Cern Virtual Machine File System, or CVMFS\cite{CVMFS}.
+The solution we opted for is CERN Virtual Machine File System, or CVMFS\cite{CVMFS}.
 
@@
 However, they did not all have the same set of system packages installed.
-Moreover, we were interested in supporting legacy clusters that were running CentOS 6, and virtual machines launched by users on one of several Compute Canada cloud platforms.
-In order to use the unified software environment on virtual machines and multiple clusters, we needed a compatibility layer.
-The solution we deployed is Nix \cite{Nix}, although we are experimenting with Gentoo Prefix \cite{Gentoo} as a replacement.
-We will discuss these options in Section~\ref{sec:Compatibility_layer}.
+Moreover, we were interested in supporting legacy clusters
+that were running CentOS 6, and virtual machines launched by users on one of several Compute Canada cloud platforms.
+In order to use the unified software environment on all of these systems, we needed a compatibility layer.
+The solution we deployed is Nix \cite{Nix}, although we are experimenting with Gentoo Prefix \cite{Gentoo} as a replacement. We will discuss these options in Section~\ref{sec:Compatibility_layer}.
 
@@
 This means that automation is needed.
-The solution identified for the compatibility layer already addresses this issue.
-However, these solutions mostly target basic Linux packages.
-For scientific packages, we also wanted to ensure that the software packages were optimized for specific hardware architectures.
-Two tools with a large set of existing recipes were known to us at the time: Spack\cite{Spack} and EasyBuild \cite{EasyBuild2012,EasyBuild2014,EasyBuild2016}.
-We will discuss why we chose EasyBuild in Section~\ref{sub:Compiled_software_packages}.
+Nix already addresses this issue; however it mostly targets basic Linux packages.
+For scientific packages, we also wanted to ensure that the software packages were optimized for specific hardware architectures, which Nix does not.
+The tool we chose for this task is EasyBuild~\cite{EasyBuild2012,EasyBuild2014,EasyBuild2016}, which we will discuss in Section~\ref{sub:Compiled_software_packages}.
 
@@
 All of our clusters were using a system of environment modules \cite{Modules1991,Modules1996}.
-However, the traditional systems of modules become unwieldy to handle at the scale of thousands of installed modules.
+However, the traditional module systems become unwieldy to handle at the scale of thousands of installed modules.
 Using a hierarchical module tree, which is possible in a newer implementation of environment modules called Lmod \cite{Lmod}, helps address this problem.
@@
 \label{sec:Distribution_mechanism}
-Installing software packages on a single cluster can be done through a shared file system, such as NFS, Lustre, or GPFS.
-Anyone who has been working in the field of high performance computing for a number of years knows that parallel file systems are often a bottleneck, especially for the IO pattern caused by the action of launching applications.
-In addition to that problem, this solution is also a single point of failure, since any failure of that shared file system will block the vast majority of jobs.
-Another solution for a single cluster is to install binary packages on each node, using a mechanism such as RPM.
-This solution can however lead to diverging configurations on different nodes, and require an additional layer to keep the configurations synchronized.
+A shared file system such as NFS, Lustre, or GPFS, can provide access to software packages on a single cluster.
+However, experience in the field of high performance computing shows that parallel file systems are often challenging to operate at scale in a reliable and performant manner, especially for the high I/O operations per seconds caused by applications startup.
+In addition, a shared file system is a single point of failure for a cluster, which nearly all jobs rely on in order to run.
+An alternative for a single cluster is to install binary packages on each node, using a mechanism such as RPM.
+However, this can result in divergent configurations across nodes, may be restricted by available space on the local disk, and requires an additional configuration management layer to keep the packages synchronized.
 
-Adding multiple clusters to the problem poses a second synchronization problem.
-To keep clusters synchronized, there are two strategies : pushing changes to the clusters, or pulling changes from a central location.
-Both strategies could be implemented with a tool such as {\it rsync}.
-However, synchronizing directories with over 40 million files and  directories takes a few hours, even with a tool as efficient as {\it rsync}.
+For a national platform spanning multiple clusters, latency precludes a parallel file system from consideration, and the synchronization problem is compounded.
+The clusters could be synchronized by pulling or pushing changes with a tool such as {\it rsync}, but this would not be scalable; even a tool as efficient as {\it rsync} takes several hours to synchronize over 40 million files and directories.
+Moreover, this method of copying files is not transactional or atomic, so an additional indirection layer would be required to provide a complete solution.
 
-We instead chose to use CERN Virtual Machine File System (CVMFS) \cite{CVMFS}.
-CVMFS is a file system initially designed to distribute software packages to an array of clusters geographically distributed around the world.
-It is a resilient file system with no single point of failure, and it delivers files to the clients in a pull fashion: each client pulls file from it as needed.
-Multiple cache layers, at the level of each node and each cluster, keep network traffic to the central repository to a minimum.
+We therefore chose to use CERN Virtual Machine File System (CVMFS) \cite{CVMFS}.
+CVMFS is a file system designed to distribute software packages to clusters at a global scale with no single point of failure.
+Clients access files by pulling content on demand, using multiple levels of caching, to minimize network traffic and storage use and ensure responsive and scalable performance.
 
@@
 CVMFS is designed similarly to the Network Time Protocol stratum model.
-There is one source of truth called the stratum 0, where any injection of new software is done.
-The stratum 0 is never accessed directly by users or even by the clusters.
+There is one source of truth called the stratum 0, where software is injected.
+The stratum 0 is never accessed directly by clients.
 If it fails, the only immediate consequence is that injection of new files is not possible until it is back online.
 
-Snapshots of the stratum 0 are kept for a while, such that it is possible to go back to a previous version if needed.
-Files in the repository are compressed, and their hash is computed to allow space saving by avoiding duplication.
-Hashesalso allows CVMFS clients to quickly invalidate their local cache as needed.
+Snapshots of the stratum 0 are kept for a while, such that it is possible to roll back to a previous version if needed.
+Files in the repository are compressed, and their hash is computed to allow space saving by deduplicating objects.
+Hashes also allow CVMFS clients to verify and invalidate their local cache quickly when contents is changed.
 
 The stratum 0 is replicated to multiple strata 1 servers that are the access points of the clusters to the software environment.
-They provide redundancy and avoid single points of failure.
+They provide redundancy and prevent single points of failure.
 Files are served to the clusters through HTTP and can be cached locally with multiple proxy servers, protecting against an outage of the wide area network.
@@
 Isolating the software environment from the operating system requires a compatibility layer containing almost everything that an OS would contain.
-From this layer, we exclude the Linux kernel and its modules, hardware drivers as well as privileged binaries.
+From this layer, we exclude the Linux kernel and its modules, hardware drivers, as well as privileged binaries.
 
@@
 Reproducibility being one the core principles in Nix, it is designed to depend as little as possible on the underlying operating system.
-Every software package installed through nix resides in a path calculated with the hash of the recipe to build it.
+Every software package installed through Nix resides in a path calculated with the hash of the recipe to build it.
 This ensures very strong reproducibility, but it also makes updating a package in place impossible.
@@
 Gentoo Prefix~\cite{Gentoo} is an alternative solution to provide a compatibility layer.
-It allows installation of most packages that make a typical Linux inside a non-default location.
+It allows installation of most packages that form a typical Linux inside a non-default location.
 Contrary to Nix however, packages are installed in a standard directory structure.
@@
 Since their performance is critical, they can not be installed in the compatibility layer, as they need to be optimized for specific CPU instructions.
-To install them, we use the EasyBuild~\cite{EasyBuild2012,EasyBuild2014,EasyBuild2016} package manager.
+To install them, we use the EasyBuild~\cite{EasyBuild2012,EasyBuild2014,EasyBuild2016} software installation tool.
 
@@
 The framework supports multiple standard and custom naming schemes.
-We use a version of a hierarchical naming scheme, in which modules are organized in a tree structure, hiding those that are incompatible currently loaded modules.
-This limits the list of modules which could otherwise be very long.
+We use a version of a hierarchical naming scheme, in which modules are organized in a tree structure, hiding those that are incompatible with currently loaded modules.
+This limits the list of visible modules which could otherwise be very extensive.
 This can however throw off some users, since they don’t automatically see all available modules.
-Lmod~\cite{Lmod} alleviates this issue by providing a command to crawl the full module tree and instructs users on how to load the module they want.
+Lmod~\cite{Lmod} alleviates this issue by providing a separate command to crawl the full module tree and instructs users on how to load the module they want.
 
@@
 regarding EasyBuild
+    \item All of the EasyBuild maintainers and contributors, without which EasyBuild would not have nearly as many supported recipes.
 	\item Kuang Chung Chen and Bryan Caron, from McGill University, for initiating the groundwork that lead to this

### 2019-02-20 extended-abstract.tex
Latest suggestions from Ryan and Bart

@@
 % The "title" command has an optional parameter, allowing the author to define a "short title" to be used in page headers.
-\title{Providing A Unified Software Environment for Canada’s National Advanced Computing Centers}
+\title{Providing a Unified Software Environment for Canada’s National Advanced Computing Centers}
 
@@
 In this paper, we present the solution that we created, which has allowed Compute Canada to serve the needs of over 10,000 researchers across the country.
-It is used on over 20 different clusters with heterogeneous configurations, from processor architectures as old as Intel Nehalem or AMD Opteron all the way up to Intel Skylake, with and without GPUs, with InfiniBand, Ethernet or OmniPath as the network fabric, and with Slurm or Torque/Moab as the scheduler.
-It presents a unified software environment to the users, providing over 600 different scientific applications that are available in over 4,000 different combinations of version, compiler and CPU architecture.
+This solution is used on over 20 different clusters with heterogeneous configurations, on processor architectures ranging from AMD's 2010 Magny-Cours to Intel's 2017 Skylake SP, with or without GPUs, with InfiniBand, Ethernet or OmniPath as the network fabric, and with Slurm or Torque/Moab as the scheduler.
+This stack provides a unified software environment to users, providing over 600 different scientific applications that are available in over 4,000 different combinations of version, compiler and CPU architecture.
 \end{abstract}
@@
 Nix already addresses this issue; however it mostly targets basic Linux packages.
-For scientific packages, we also wanted to ensure that the software packages were optimized for specific hardware architectures, which Nix does not.
+For scientific packages, we also wanted to ensure that the software packages were optimized for specific hardware architectures, which Nix can not easily do.
 The tool we chose for this task is EasyBuild~\cite{EasyBuild2012,EasyBuild2014,EasyBuild2016}, which we will discuss in Section~\ref{sub:Compiled_software_packages}.
@@
 A shared file system such as NFS, Lustre, or GPFS, can provide access to software packages on a single cluster.
-However, experience in the field of high performance computing shows that parallel file systems are often challenging to operate at scale in a reliable and performant manner, especially for the high I/O operations per seconds caused by applications startup.
+However, experience in the field of high performance computing shows that parallel file systems are often challenging to operate at scale in a reliable and performant manner, especially for the high I/O operations per second caused by applications startup.
 In addition, a shared file system is a single point of failure for a cluster, which nearly all jobs rely on in order to run.
 An alternative for a single cluster is to install binary packages on each node, using a mechanism such as RPM.
-However, this can result in divergent configurations across nodes, may be restricted by available space on the local disk, and requires an additional configuration management layer to keep the packages synchronized.
+However, this can result in divergent configurations across nodes, may be limited by available space on the local disk, and requires an additional configuration management layer to keep the packages synchronized.
 
@@
 Snapshots of the stratum 0 are kept for a while, such that it is possible to roll back to a previous version if needed.
-Files in the repository are compressed, and their hash is computed to allow space saving by deduplicating objects.
-Hashes also allow CVMFS clients to verify and invalidate their local cache quickly when contents is changed.
+Files in the repository are compressed, and their hash is computed to save space by deduplicating objects.
+Hashes also allow CVMFS clients to quickly verify and invalidate their local caches when content is changed.
 
-The stratum 0 is replicated to multiple strata 1 servers that are the access points of the clusters to the software environment.
-They provide redundancy and prevent single points of failure.
-Files are served to the clusters through HTTP and can be cached locally with multiple proxy servers, protecting against an outage of the wide area network.
-The closest layer of cache can be configured either to use a local disk on the compute nodes, or on a shared file system on in a diskless environment.
+The stratum 0 is replicated to multiple strata 1 servers, which serve content to clients and provide redundancy and load balancing.
+Files are served to the clusters via HTTP, and each cluster has local caching proxy servers to reduce the load placed on the stratum 1 servers by its worker nodes mitigating the impact of a wide area network outage.
+The closest layer of cache can be configured to use a local disk on the compute nodes, or a shared file system in a diskless environment.
 
@@
 \label{sec:Compatibility_layer}
-Isolating the software environment from the operating system requires a compatibility layer containing almost everything that an OS would contain.
-From this layer, we exclude the Linux kernel and its modules, hardware drivers, as well as privileged binaries.
+Isolating the software environment from the operating system requires a compatibility layer containing almost everything that an OS would provide.
+From this layer, we exclude the Linux kernel and its modules, hardware drivers, and privileged binaries.
 
@@
 Nix \cite{Nix} is both a package manager and an operating system (NixOS).
-Reproducibility being one the core principles in Nix, it is designed to depend as little as possible on the underlying operating system.
+One of the core principles in Nix is reproducibility, so it is designed to depend as little as possible on the underlying operating system.
 Every software package installed through Nix resides in a path calculated with the hash of the recipe to build it.
@@
 While reproducibility is an important consideration in advanced research computing, bugs sometimes need to be fixed, which is difficult due to Nix' design.
-At the time of writing this paper, we are still using Nix as a compatibility layer, but we are investigating a switch to Gentoo Prefix described next.
+This difficulty arises in particular because our Nix-installed packages are linked into software that is not installed via Nix.
+At the time of writing this paper, we are still using Nix as a compatibility layer, but we are investigating a switch to Gentoo Prefix.
 
@@
 Gentoo Prefix~\cite{Gentoo} is an alternative solution to provide a compatibility layer.
-It allows installation of most packages that form a typical Linux inside a non-default location.
+It allows installation of most packages that form a typical Linux operating system in a non-default location.
 Contrary to Nix however, packages are installed in a standard directory structure.
@@
 
-EasyBuild originated from University of Ghent, but has now over 200 contributors from all around the world.
+EasyBuild originated from the University of Ghent, but now has over 200 contributors from around the world.
 The framework is based on scripted recipes to install packages, usually from source code.
@@
 \label{sub:Handling_multiple_CPU_architectures}
-Handling multiple CPU architectures is done by installing applications in architecture dependent paths and module trees.
+Handling multiple CPU architectures is done by installing applications in architecture-dependent paths and module trees.
 We also have a wide selection of software packages that are installed in an architecture-independent fashion.
-This is the case, for example, for binary packages, packages that are not performance critical, or packages that consists of source files only.
+This is the case, for example, for binary packages, packages that are not performance critical, or packages that consist of source files only.
 The same recipe is typically used for installing the same package for multiple architectures.
@@
 Wheels are binary packages that can be installed in a matter of seconds, similar to CentOS RPM packages.
-Since they are built by our team, and linked against our optimize software stack, we ensure that they are both easy to installed and offer good performance.
+Since they are built by our team, and linked against our optimized software stack, we ensure that they are easy to install and offer good performance.
 
@@
 Because of the portability of our software stack and the distribution mechanism, it can be used on any cluster, or any virtual machine on a cloud.
-For example, for training events, we have been using it to provision small sized replicas of Compute Canada clusters in short-lived virtual clusters deployed on our OpenStack platform.
+For example, for training events, we have been using it to provision small replicas of Compute Canada clusters in short-lived virtual clusters deployed on our OpenStack platform.
 We have also used this portable software environment to expose a cluster environment directly through a browser, via JupyterHub.
@@
     \item All of the EasyBuild maintainers and contributors, without which EasyBuild would not have nearly as many supported recipes.
-	\item Kuang Chung Chen and Bryan Caron, from McGill University, for initiating the groundwork that lead to this
-project
+	\item Kuang Chung Chen and Bryan Caron, from McGill University, for initiating the groundwork in Compute Canada that lead to this project
 	\item Dave Dykstra, Jakob Blomer, and the CernVM project team for support and development of CVMFS

### 2019-02-25 paper.tex
Added paper.tex
Changes done by Ryan.

--- 2019-02-20 extended-abstract.tex
@@
 % the first part of the formatted document.
-%\maketitle
+\maketitle
 
@@
 Moreover, we were interested in supporting legacy clusters
-that were running CentOS 6, and virtual machines launched by users on one of several Compute Canada cloud platforms.
+that were running CentOS 6, and virtual machines launched by users on several Compute Canada cloud platforms.
 In order to use the unified software environment on all of these systems, we needed a compatibility layer.
@@
 Nix already addresses this issue; however it mostly targets basic Linux packages.
-For scientific packages, we also wanted to ensure that the software packages were optimized for specific hardware architectures, which Nix can not easily do.
+For scientific packages, we also wanted to ensure that the software packages were optimized for specific hardware architectures, which Nix cannot do easily.
 The tool we chose for this task is EasyBuild~\cite{EasyBuild2012,EasyBuild2014,EasyBuild2016}, which we will discuss in Section~\ref{sub:Compiled_software_packages}.
@@
 All of our clusters were using a system of environment modules \cite{Modules1991,Modules1996}.
-However, the traditional module systems become unwieldy to handle at the scale of thousands of installed modules.
+However, the traditional module systems become unwieldy at the scale of thousands of installed modules.
 Using a hierarchical module tree, which is possible in a newer implementation of environment modules called Lmod \cite{Lmod}, helps address this problem.
@@
 A shared file system such as NFS, Lustre, or GPFS, can provide access to software packages on a single cluster.
-However, experience in the field of high performance computing shows that parallel file systems are often challenging to operate at scale in a reliable and performant manner, especially for the high I/O operations per second caused by applications startup.
+However, experience in the field of high performance computing shows that parallel file systems are often challenging to operate at scale in a reliable and performant manner, especially for the high I/O operations per second caused by application startups.
 In addition, a shared file system is a single point of failure for a cluster, which nearly all jobs rely on in order to run.
@@
 
-Snapshots of the stratum 0 are kept for a while, such that it is possible to roll back to a previous version if needed.
+Snapshots of the stratum 0 are kept for a while, making it possible to roll back to a previous version if needed.
 Files in the repository are compressed, and their hash is computed to save space by deduplicating objects.
@@
 The stratum 0 is replicated to multiple strata 1 servers, which serve content to clients and provide redundancy and load balancing.
-Files are served to the clusters via HTTP, and each cluster has local caching proxy servers to reduce the load placed on the stratum 1 servers by its worker nodes mitigating the impact of a wide area network outage.
+Files are served to the clusters via HTTP, and each cluster has local caching proxy servers to reduce the load placed on the stratum 1 servers by its worker nodes, and mitigate the impact of a wide area network outage.
 The closest layer of cache can be configured to use a local disk on the compute nodes, or a shared file system in a diskless environment.
@@
 Isolating the software environment from the operating system requires a compatibility layer containing almost everything that an OS would provide.
-From this layer, we exclude the Linux kernel and its modules, hardware drivers, and privileged binaries.
+We exclude the Linux kernel and its modules, hardware drivers, and privileged binaries from this layer.
 

### 2019-03-21 paper.tex
Added details about Compute Canada's CVMFS repositories

@@
 
+We will describe the generic structure of a CVMFS deployment in Section~\ref{sub:CVMFS_structure}, and describe the deployment specific to Compute Canada in Section~\ref{sub:CVMFS_in_CC}
+
 \subsection{CVMFS structure}
@@
 The closest layer of cache can be configured to use a local disk on the compute nodes, or a shared file system in a diskless environment.
+
+\subsection{Deployment of CVMFS in Compute Canada}
+\label{sub:CVMFS_in_CC}
+For Compute Canada's needs, we have deployed three CVMFS repositories, each composed of a stratum 0 and one or multiple strata 1.
+
+First, we have a development repository.
+This repository is used to deploy software from our build infrastructure onto the clusters without exposing it to end users.
+This allows us test newly installed software packages on our clusters without impacting production.
+This minimizes the odds of a faulty package being pushed to production.
+This is especially important when a bug is found in one of the software packages and we need to update it.
+This also happens for example if we need to make minor changes to an MPI implementation in order to correctly support a new cluster with a different networking fabric.
+
+We then have a production repository.
+By design, software packages are always copied onto the production repository from the development repository, and never from the build infrastructure directly.
+This workflow also minimizes the odds of a faulty software making its way to the end users.
+This repository holds all of the software environment which is visible to the end users on all of our clusters.
+Once a piece of software is deployed onto this repository, it becomes visible on all clusters within 30 minutes.
+
+Both repositories above are available publicly, and can be mounted on any cluster or virtual machine in the cloud.
+We also have a third repository which is used to host software packages that are restricted to Compute Canada due to license limitations.
+We work with software vendors and developers in order to get the authorization to install licensed and commercial software packages in this repository.
+The most common licensing model that we use is a BYOL (bring your own license) model, in which we provide the software package, but end users provide a license server to connect to.
+Another model that is used is to control access to a software package via POSIX groups.
+By default, CVMFS clients do not honour POSIX group and filesystem permissions, but they can be configured to do so.
+Since this kind of restriction requires a specific LDAP and a specific configuration, this repository is restricted and can only be made available to clusters that are run by Compute Canada teams.
+
+One final use case for CVMFS is to hold and distribute datasets.
+However, default CVMFS configuration is optimized for a large number of small files, rather than a small number of large files.
+This means that a special configuration must be done for this repository.
+At the time of writing this paper, this use case is still experimental within Compute Canada, but other groups are known to use CVMFS in this way.
 

### 2019-03-21 paper.tex
Added details about Nix

@@
 \label{sub:Nix}
-Nix \cite{Nix} is both a package manager and an operating system (NixOS).
+Nix~\cite{Nix} is both a package manager and an operating system (NixOS).
 One of the core principles in Nix is reproducibility, so it is designed to depend as little as possible on the underlying operating system.
-Every software package installed through Nix resides in a path calculated with the hash of the recipe to build it.
-This ensures very strong reproducibility, but it also makes updating a package in place impossible.
-While reproducibility is an important consideration in advanced research computing, bugs sometimes need to be fixed, which is difficult due to Nix' design.
-This difficulty arises in particular because our Nix-installed packages are linked into software that is not installed via Nix.
-At the time of writing this paper, we are still using Nix as a compatibility layer, but we are investigating a switch to Gentoo Prefix.
+
+The installation path of each software package installed through Nix resides in a path called the Nix store.
+This stores contains a single directory for each version of each software package installed through Nix.
+These directories are named after a hash calculated from the installation recipe and its dependencies.
+This creates a very large number of subdirectories, many of which contain almost identical files.
+
+In addition to the store, Nix creates directories that are called profiles.
+These profile directories contain a farm of symbolic links that point to files in the Nix store.
+For example, it will contain a {\it bin} directory, which will contain symbolic links to typical operating system commands such as {\it ls} or {\it gcc}, which will reside in different paths in the store.
+It will also contain {\it lib} and {\it include} directories which will contain symbolic links to libraries and header files located in the store.
+
+Profiles are the paths that are exposed to the end users, which is done by adding paths to their {\it PATH}, {\it CPATH} and other similar environment variables.
+Each profile ever created is kept, but users are only exposed to the most recent one.
+Each profile represents the complete state of the user-exposed environment at a given time in the past.
+
+Nix has a mechanism for garbage collection, which can be used to delete packages that are no longer used in the current profile.
+It also has a mechanism to resolve the complete list of files from the Nix store that are required to compose a given profile.
+Because Nix generates so many files, we use this mechanism to deploy on CVMFS only those files that are used in the current profile.
+
+This works well until a software exposes paths to the Nix store directly, instead of the paths to Nix profiles.
+We call this a Nix store {\it leak}.
+This happens for example with Python when using virtual environments.
+Virtual environments in Python copy the {\it python} executable to a secondary location, outside of Nix.
+However, all binaries installed by Nix use {\it RPATH} for shared object resolutions in the store directly.
+This means that copying a binary outside of Nix creates a dependency between a binary that now resides outside of the store, and the store itself.
+Because Nix is unaware of that dependency, any update of one of the dependencies followed by garbage collection or selective deployment of files will end up breaking this binary.
+We have encountered this issue with Python, Perl and Qt, which we have therefore moved outside of Nix and installed them through EasyBuild instead.
+
+Another issue that we have encountered is that many Nix recipes call for installing sub-packages of a package in separate directories.
+For example, for Qt, {\it qmake} and {\it OTHER COMMAND} reside in different directories in the Nix store.
+This sometimes cause issues when installing a dependent package which assumes that its dependencies are installed in a more traditional fashion, with all of the binaries in the same {\it bin} folder, alongside with headers in the {\it include} folder, and libraries in the {\it lib} folder.
+This has mostly been an issue for Qt.
 
@@
 It allows installation of most packages that form a typical Linux operating system in a non-default location.
-Contrary to Nix however, packages are installed in a standard directory structure.
+Contrary to Nix however, packages are installed in a standard directory structure ({\it bin}, {\it lib}, {\it include}, {\it share}).
 This is helpful when updates are required, and minimizes problems with applications that expect a standard directory structure.

### 2019-03-21 paper.tex
Added details about EasyBuild

@@
 EasyBuild originated from the University of Ghent, but now has over 200 contributors from around the world.
-The framework is based on scripted recipes to install packages, usually from source code.
+The framework is based on scripted recipes (in Python) to install packages, usually from source code.
 It is designed to create optimized binaries, and eases recompilation of multiple versions of a given application using many combinations of compilers or libraries.
+EasyBuild is composed of three parts: the framework, the EasyBlocks, and the EasyConfigs.
+
+The framework contains all of the core code which drives the logic of installing a software.
+It drives all of the steps necessary to install a software: parsing the configuration options, downloading source files, patching them, configuring, building and installing the software, as well as performing post-installation steps such as sanity checks, generating a module or running custom commands.
+
+Written in Python, EasyBlocks define what needs to be done at each step for a given software package.
+There are generick EasyBlocks, such as {\it ConfigureMake} or {\it CMakeMake}, as well as custom EasyBlocks for packages that have more complicated installation procedures, such as OpenFOAM or PETSc.
+
+EasyConfigs files parametrize the execution of the code from the EasyBlocks.
+They define the software name and version, the URL from which the source files can be downloaded as well as options to be used for the various steps.
+
+In most cases, there is already an EasyBlock for a given software package, or it can be installed easily with one of the generic EasyBlocks.
+In these cases, writing a new recipe is as simple as editing a configuration file.
+EasyBuild also supports command line options to reuse and existing EasyConfig file, while changing the software version or the toolchain being used.
+
+Toolchains are a core concept in EasyBuild.
+They represent a set or subset of a compiler, and MPI implementation, mathematical libraries and CUDA.
+Having the ability to generate multiple builds of a given software simply by providing a command line argument to change the toolchain has proven unvaluable to our workflow, and has allowed us to recompile hundreds of packages in matter of days when deploying a new version of compiler or a cluster with a new CPU architecture.
 

### 2019-03-21 paper.tex
added references to figures to be added and a description of the layered approach for System/Nix/EasyBuild

@@
 \label{sub:CVMFS_structure}
-CVMFS is designed similarly to the Network Time Protocol stratum model.
+CVMFS is designed similarly to the Network Time Protocol stratum model, as illustrated in Figure~\ref{fig:CVMFS_structure}.
 There is one source of truth called the stratum 0, where software is injected.
@@
 
+\textbf{TODO: Insert figure representing the CVMFS structure of strata. }
+\label{fig:CVMFS_structure}
+
 \subsection{Deployment of CVMFS in Compute Canada}
@@
 For Compute Canada's needs, we have deployed three CVMFS repositories, each composed of a stratum 0 and one or multiple strata 1.
+This allows us to follow a workflow that includes testing before exposing software packages to user, as illustrated in Fig~\ref{fig:Workflow}.
 
@@
 At the time of writing this paper, this use case is still experimental within Compute Canada, but other groups are known to use CVMFS in this way.
+
+\textbf{TODO: INSERT FIGURE representing our workflow (build => test locally => dev => test on cluster => prod) }
+\label{fig:Workflow}
+
+\section{A layered approach}
+\label{sec:Layered_environment}
+To install software packages for our common environment, we follow a layered approach, as illustrated in Table~\ref{tab:layers}.
+First, we require a basic Linux installation.
+Any version of Linux with a kernel 2.6.32 or more recent should work.
+This layer is managed by the system administrators, and should contain any driver, such as GPU drivers or drivers for the network fabric, kernel modules, as well as privileged binaries.
+
+The second layer is what we call our compatibility layer.
+It contains almost all packages that one would typically install with the Linux distribution's package manager, such as {\it yum} or {\it apt-get}.
+
+The third layer is the scientific application layer.
+It contains mostly scientific or performance critical applications, including for example MPI implementations, CUDA, mathematical libraries, and domain specific applications.
+
+\textbf{TODO: INSERT FIGURE/TABLE WITH OS layer, Nix layer, EasyBuild layer}
+\label{tab:layers}
 

### 2019-03-21 paper.tex
correction of typos

@@
 
-Written in Python, EasyBlocks define what needs to be done at each step for a given software package.
+EasyBlocks are Python classes that define the steps required to install a given software package.
 There are generick EasyBlocks, such as {\it ConfigureMake} or {\it CMakeMake}, as well as custom EasyBlocks for packages that have more complicated installation procedures, such as OpenFOAM or PETSc.
@@
 Toolchains are a core concept in EasyBuild.
-They represent a set or subset of a compiler, and MPI implementation, mathematical libraries and CUDA.
+They represent a set or subset of a compiler, an MPI implementation, mathematical libraries and CUDA.
 Having the ability to generate multiple builds of a given software simply by providing a command line argument to change the toolchain has proven unvaluable to our workflow, and has allowed us to recompile hundreds of packages in matter of days when deploying a new version of compiler or a cluster with a new CPU architecture.

### 2019-03-21 paper.tex
Added description of how to make it available on a computer. Added subsection about logging and documentation. Added conclusion

@@
 
-\section{Other benefits}
-\label{sec:Other_benefits}
+\subsection{Documentation of supported packages}
+\label{sub:Documentation}
+Every software package that is installed through EasyBuild get automatically listed on our documentation website, at \\
+\url{https://docs.computecanada.ca/wiki/Available_software}
+A similar list of available Python wheels is automatically generated at \\
+\url{https://docs.computecanada.ca/wiki/Available_wheels}
+
+\subsection{Logging of module usage}
+\label{sub:Logging}
+Our module system will automatically log (through {\it logger}) each module that is loaded.
+The information is logged in an easily parsable list of key-value pairs, which can then be aggregated into your favorite monitoring tool such as Elastic Search.
+
+\section{Making the software environment available on your cluster or computer}
+\label{sec:Making_it_available}
+Making our software environment available on a single computer is as simple as installing the CVMFS client and configuring repositories manually or installing our RPM
+\textbf{TODO: Document our RPM}
+
+After CVMFS is configured, one can simply execute the command below in order to obtain Compute Canada's software environment \\
+\texttt{source /cvmfs/soft.computecanada.ca/config/profile/bash} \\
+
+By default, our default set of modules will try to detect the most recent set of instructions that is supported by the CPU amongst {\it SSE3}, {\it AVX}, {\it AVX2} and {\it AVX512}.
+It will also try to detect whether the hardware has a {\it Mellanox InfiniBand}, {\it Intel OmniPath}, or an Ethernet network fabric.
+If these are not detected correctly, or if you have a cluster made up of mixed hardware, you can overwrite these defaults by defining the {\verb!RSNT_ARCH!} or {\verb!RSNT_INTERCONNECT!} environment variables before executing the command above.
+
+Note that the first time a command is ran, it will take some longer time to start the command, as the CVMFS client is fetching the files that are needed to run the command from our stratum 1 servers.
+Following executions of the same command will however be fast as files will have been cached locally.
+
+Also note that the command above will \textbf{not} import our environment for any user ID lower than 1000, which are typically privileged users on a Linux system.
+In order to force the command above to execute nevertheless, you can create a file named \verb!.force_cc_cvmfs!
+in the user's \verb!$HOME!
+directory.
+
+Making our software environment available on a cluster requires the additional step of configuring one or multiple HTTP proxies.
+This is required to avoid overloading the stratum-1 server by having a large number of nodes all fetching the same files at the same time.
+If you are interested in testing or making our software environment available to your users on your cluster, we ask that you first contact us.
+Note that this software stack comes with absolutely no warranty.
+
+\section{Other benefits and conclusion}
+\label{sec:Conclusion}
+We have built a software environment that is portable, is optimized for multiple CPU architectures and network interconnect, and can be deployed in a matter of minutes on a single node, and barely longer for a whole cluster.
+The design of the infrastructure guarantees that the software environment seen by our end users is the same across all of our clusters, and will not diverge over time.
+This makes for a better user experience, as it becomes much easier to migrate between two clusters.
+It also reduces the workload on our staff by unifying the installation process for all clusters, and makes supporting our users easier because all clusters work the same way.
+
 Because of the portability of our software stack and the distribution mechanism, it can be used on any cluster, or any virtual machine on a cloud.

### 2019-03-26 paper.tex
Added two figures

@@
 
-\textbf{TODO: Insert figure representing the CVMFS structure of strata. }
+\begin{figure}
+  \includegraphics[width=0.45\textwidth]{CVMFS-structure.pdf}
+  \caption{Diagram of the CVMFS structure. Arrows represent the flow of data. Data is pushed into the stratum-0, but is then fetched by the following layers.}
   \label{fig:CVMFS_structure}
+\end{figure}
 
@@
 
-\textbf{TODO: INSERT FIGURE representing our workflow (build => test locally => dev => test on cluster => prod) }
+\begin{figure}
+  \includegraphics[width=0.45\textwidth]{Software-installation-workflow.pdf}
+  \caption{Workflow followed by our software installers. At each step, there is a checkpoint to test if the software package works properly.}
   \label{fig:Workflow}
+\end{figure}
 

### 2019-03-26 paper.tex
Added last figure

@@
 
-\textbf{TODO: INSERT FIGURE/TABLE WITH OS layer, Nix layer, EasyBuild layer}
-\label{tab:layers}
+\begin{figure*}
+  \includegraphics[width=0.7\textwidth]{layers.png}
+  \caption{Layered structure of our software stack.
+The bottom layer (red) is entirely managed by the system administrators of each cluster.
+The second layer (grey) contains libraries that can be installed centrally, through Nix, but can be overridden locally.
+The third layer (yellow) contains everything that one would typically find in the Linux user space.
+The fourth layer (green) contains packages that you would typically find in an operating system, but for which multiple versions are required.
+They are installed through Nix, but wrapped into a module by EasyBuild.
+The top layer (blue) contains packages that are not typically installed in a Linux distribution, and are installed via EasyBuild.}
+  \label{fig:layers}
+\end{figure*}
 

### 2019-04-05 paper.tex
Corrections

@@
 }
-\author{Bart Oldeman}
+\author{Bart E. Oldeman}
 \email{bart.oldeman@calculquebec.ca}
@@
 \label{sec:Distribution_mechanism}
-A shared file system such as NFS, Lustre, or GPFS, can provide access to software packages on a single cluster.
+A shared file system such as NFS, Lustre, or GPFS can provide access to software packages on a single cluster.
 However, experience in the field of high performance computing shows that parallel file systems are often challenging to operate at scale in a reliable and performant manner, especially for the high I/O operations per second caused by application startups.
@@
 First, we require a basic Linux installation.
-Any version of Linux with a kernel 2.6.32 or more recent should work.
-This layer is managed by the system administrators, and should contain any driver, such as GPU drivers or drivers for the network fabric, kernel modules, as well as privileged binaries.
+Any version of Linux with a kernel 2.6.32 (as used by RHEL 6 and its derivatives) or more recent should work.
+This layer is managed by the system administrators, and contains all drivers, typically provided by kernel modules and closely associated libraries, such as GPU and network drivers, as well as privileged binaries.
 
@@
 The third layer is the scientific application layer.
-It contains mostly scientific or performance critical applications, including for example MPI implementations, CUDA, mathematical libraries, and domain specific applications.
+It contains mostly scientific or performance critical applications, including for example MPI implementations, the CUDA toolkit, mathematical libraries, and domain specific applications.
 
@@
 The top layer (blue) contains packages that are not typically installed in a Linux distribution, and are installed via EasyBuild.}
-  \label{fig:layers}
+  \label{tab:layers}
 \end{figure*}
@@
 It is designed to create optimized binaries, and eases recompilation of multiple versions of a given application using many combinations of compilers or libraries.
-EasyBuild is composed of three parts: the framework, the EasyBlocks, and the EasyConfigs.
+EasyBuild is composed of three parts: the framework, easyblocks, and easyconfig files (also known as easyconfigs).
 
@@
 
-EasyBlocks are Python classes that define the steps required to install a given software package.
-There are generick EasyBlocks, such as {\it ConfigureMake} or {\it CMakeMake}, as well as custom EasyBlocks for packages that have more complicated installation procedures, such as OpenFOAM or PETSc.
+Easyblocks are Python classes that define the steps required to install a given software package.
+There are generic easyblocks, such as {\it ConfigureMake} or {\it CMakeMake}, as well as custom easyblocks for packages that have more complicated installation procedures, such as OpenFOAM or PETSc.
 
-EasyConfigs files parametrize the execution of the code from the EasyBlocks.
+Easyconfig files parametrize the execution of the code from the easyblocks.
 They define the software name and version, the URL from which the source files can be downloaded as well as options to be used for the various steps.
 
-In most cases, there is already an EasyBlock for a given software package, or it can be installed easily with one of the generic EasyBlocks.
+In most cases, there is already an easyblock for a given software package, or it can be installed easily with one of the generic easyblocks.
 In these cases, writing a new recipe is as simple as editing a configuration file.
-EasyBuild also supports command line options to reuse and existing EasyConfig file, while changing the software version or the toolchain being used.
+EasyBuild also supports command line options to reuse an existing easyconfig file, while changing the software version or the toolchain being used.
 
@@
 They represent a set or subset of a compiler, an MPI implementation, mathematical libraries and CUDA.
-Having the ability to generate multiple builds of a given software simply by providing a command line argument to change the toolchain has proven unvaluable to our workflow, and has allowed us to recompile hundreds of packages in matter of days when deploying a new version of compiler or a cluster with a new CPU architecture.
+Having the ability to generate multiple builds of a given software simply by providing a command line argument to change the toolchain has proven invaluable to our workflow, and has allowed us to recompile hundreds of packages in matter of days when deploying a new version of compiler or a cluster with a new CPU architecture.
 

### 2019-04-08 extended-abstract.tex
Corrections
Integrated Ryan's comments

@@
 However, they did not all have the same set of system packages installed.
-Moreover, we were interested in supporting legacy clusters
-that were running CentOS 6, and virtual machines launched by users on one of several Compute Canada cloud platforms.
+Moreover, we were interested in supporting legacy clusters that were running CentOS 6, and virtual machines launched by users on several Compute Canada cloud platforms.
 In order to use the unified software environment on all of these systems, we needed a compatibility layer.
@@
 \label{sec:Distribution_mechanism}
-A shared file system such as NFS, Lustre, or GPFS, can provide access to software packages on a single cluster.
-However, experience in the field of high performance computing shows that parallel file systems are often challenging to operate at scale in a reliable and performant manner, especially for the high I/O operations per second caused by applications startup.
+A shared file system such as NFS, Lustre, or GPFS can provide access to software packages on a single cluster.
+However, experience in the field of high performance computing shows that parallel file systems are often challenging to operate at scale in a reliable and performant manner, especially for the high I/O operations per second caused by applications starting.
 In addition, a shared file system is a single point of failure for a cluster, which nearly all jobs rely on in order to run.
 An alternative for a single cluster is to install binary packages on each node, using a mechanism such as RPM.
-However, this can result in divergent configurations across nodes, may be limited by available space on the local disk, and requires an additional configuration management layer to keep the packages synchronized.
+However, this can result in divergent configurations across nodes, may be limited by available space on the local disks, and requires an additional configuration management layer to keep the packages synchronized.
 
@@
 
-\subsection{CVMFS structure}
+\subsection{CVMFS architecture}
 \label{sub:CVMFS_structure}
 CVMFS is designed similarly to the Network Time Protocol stratum model.
-There is one source of truth called the stratum 0, where software is injected.
+There is one source of truth, the stratum 0 server, where software is injected.
 The stratum 0 is never accessed directly by clients.
@@
 
-Snapshots of the stratum 0 are kept for a while, such that it is possible to roll back to a previous version if needed.
-Files in the repository are compressed, and their hash is computed to save space by deduplicating objects.
-Hashes also allow CVMFS clients to quickly verify and invalidate their local caches when content is changed.
+Snapshots of the software repositories on the stratum 0 are kept for a while, making it possible to roll back to a previous version if needed.
+Internally, CVMFS uses a content-addressable storage scheme, wherein files exceeding a certain size are broken into smaller objects, and each object is hashed, compressed, and stored in a location defined by its hash, resulting in natural deduplication.
+Hashes also allow CVMFS clients to quickly verify the integrity of content, and invalidate their local caches when content is changed, by means of a Merkle tree.
 
-The stratum 0 is replicated to multiple strata 1 servers, which serve content to clients and provide redundancy and load balancing.
-Files are served to the clusters via HTTP, and each cluster has local caching proxy servers to reduce the load placed on the stratum 1 servers by its worker nodes mitigating the impact of a wide area network outage.
+The stratum 0 is replicated to multiple stratum 1 servers, which serve content to clients and provide redundancy and load balancing.
+Files are served to the clusters via HTTP, and each cluster has local caching proxy servers to reduce the load placed on the stratum 1 servers by its worker nodes, which also mitigates the impact of a wide area network outage.
 The closest layer of cache can be configured to use a local disk on the compute nodes, or a shared file system in a diskless environment.
+Clients can also be configured to use two levels of cache together in a tiered manner, such as a solid state drive and a hard drive, or a hard drive and a shared file system.
 

### 2019-04-08 paper.tex
Integrated Ryan's comments

@@
 A shared file system such as NFS, Lustre, or GPFS can provide access to software packages on a single cluster.
-However, experience in the field of high performance computing shows that parallel file systems are often challenging to operate at scale in a reliable and performant manner, especially for the high I/O operations per second caused by application startups.
+However, experience in the field of high performance computing shows that parallel file systems are often challenging to operate at scale in a reliable and performant manner, especially for the high rate of I/O operations per second caused by applications starting.
 In addition, a shared file system is a single point of failure for a cluster, which nearly all jobs rely on in order to run.
 An alternative for a single cluster is to install binary packages on each node, using a mechanism such as RPM.
-However, this can result in divergent configurations across nodes, may be limited by available space on the local disk, and requires an additional configuration management layer to keep the packages synchronized.
+However, this can result in divergent configurations across nodes, may be limited by available space on the local disks, and requires an additional configuration management layer to keep the packages synchronized.
 
 For a national platform spanning multiple clusters, latency precludes a parallel file system from consideration, and the synchronization problem is compounded.
-The clusters could be synchronized by pulling or pushing changes with a tool such as {\it rsync}, but this would not be scalable; even a tool as efficient as {\it rsync} takes several hours to synchronize over 40 million files and directories.
+The clusters could be synchronized by pulling or pushing changes with a tool such as \texttt{rsync}, but this would not be scalable; even a tool as efficient as \texttt{rsync} takes several hours to synchronize over 40 million files and directories.
 Moreover, this method of copying files is not transactional or atomic, so an additional indirection layer would be required to provide a complete solution.
 
-We therefore chose to use CERN Virtual Machine File System (CVMFS) \cite{CVMFS}.
+We therefore chose to use CERN Virtual Machine File System (CVMFS) \cite{CVMFS,CVMFS2012}.
 CVMFS is a file system designed to distribute software packages to clusters at a global scale with no single point of failure.
@@
 
-We will describe the generic structure of a CVMFS deployment in Section~\ref{sub:CVMFS_structure}, and describe the deployment specific to Compute Canada in Section~\ref{sub:CVMFS_in_CC}
+We will describe the general architecture of a CVMFS deployment in Section~\ref{sub:CVMFS_structure}, and describe the deployment specific to Compute Canada in Section~\ref{sub:CVMFS_in_CC}
 
-\subsection{CVMFS structure}
+\subsection{CVMFS architecture}
 \label{sub:CVMFS_structure}
 CVMFS is designed similarly to the Network Time Protocol stratum model, as illustrated in Figure~\ref{fig:CVMFS_structure}.
-There is one source of truth called the stratum 0, where software is injected.
+There is one source of truth, the stratum 0 server, where software is injected.
 The stratum 0 is never accessed directly by clients.
@@
 
-Snapshots of the stratum 0 are kept for a while, making it possible to roll back to a previous version if needed.
-Files in the repository are compressed, and their hash is computed to save space by deduplicating objects.
-Hashes also allow CVMFS clients to quickly verify and invalidate their local caches when content is changed.
+Snapshots of the software repositories on the stratum 0 are kept for a while, making it possible to roll back to a previous version if needed.
+Internally, CVMFS uses a content-addressable storage scheme, wherein files exceeding a certain size are broken into smaller objects, and each object is hashed, compressed, and stored in a location defined by its hash, resulting in automatic deduplication.
+Hashes also allow CVMFS clients to quickly verify the integrity of content received over the network, and invalidate their local caches when content is changed, by means of a Merkle tree.
 
-The stratum 0 is replicated to multiple strata 1 servers, which serve content to clients and provide redundancy and load balancing.
-Files are served to the clusters via HTTP, and each cluster has local caching proxy servers to reduce the load placed on the stratum 1 servers by its worker nodes, and mitigate the impact of a wide area network outage.
-The closest layer of cache can be configured to use a local disk on the compute nodes, or a shared file system in a diskless environment.
+Data is replicated from the stratum 0 to multiple stratum 1 servers, which in turn serve content to clients via HTTP and provide redundancy and load balancing.
+Each cluster has local caching proxy servers to reduce the load placed on the stratum 1 servers by its compute nodes, accelerate access to any content missing from the local cache of a compute node, and mitigate the impact of a wide area network outage.
+The closest layer of cache can be configured to use a local disk on a node, or a shared file system in a diskless environment.
+Clients can also be configured to use two levels of cache together in a tiered manner, such as a solid state drive and a hard drive, or a hard drive and a shared file system.
 
@@
   \includegraphics[width=0.45\textwidth]{CVMFS-structure.pdf}
-  \caption{Diagram of the CVMFS structure. Arrows represent the flow of data. Data is pushed into the stratum-0, but is then fetched by the following layers.}
+  \caption{Diagram of the architecture of CVMFS. Arrows represent the flow of data. Data is pushed to the stratum 0, but is then fetched by the following layers.}
   \label{fig:CVMFS_structure}
@@
 \label{sub:CVMFS_in_CC}
-For Compute Canada's needs, we have deployed three CVMFS repositories, each composed of a stratum 0 and one or multiple strata 1.
-This allows us to follow a workflow that includes testing before exposing software packages to user, as illustrated in Fig~\ref{fig:Workflow}.
+For Compute Canada's needs, we deployed two CVMFS server stacks, one for development and one for production, each of which is composed of a stratum 0 server and one or more stratum 1 servers.
+This allows us to follow a workflow that includes testing software packages before exposing them to users, as illustrated in Fig~\ref{fig:Workflow}.
+In total, three software repositories are used.
 
-First, we have a development repository.
-This repository is used to deploy software from our build infrastructure onto the clusters without exposing it to end users.
-This allows us test newly installed software packages on our clusters without impacting production.
-This minimizes the odds of a faulty package being pushed to production.
-This is especially important when a bug is found in one of the software packages and we need to update it.
-This also happens for example if we need to make minor changes to an MPI implementation in order to correctly support a new cluster with a different networking fabric.
+First, we have a development repository, hosted on the development servers.
+This repository is used to deploy software from our build infrastructure to the clusters without exposing it to end users.
+This allows us to test newly installed software packages on our clusters without impacting production, minimizing the risk of a faulty package causing problems for users.
+This is especially important, for example, when testing updates to software packages to fix bugs, or to make minor changes to an MPI implementation in order to correctly support a new cluster with a different networking fabric.
 
-We then have a production repository.
-By design, software packages are always copied onto the production repository from the development repository, and never from the build infrastructure directly.
-This workflow also minimizes the odds of a faulty software making its way to the end users.
+We also have a production repository, hosted on the production servers.
+By design, software packages are always copied to the production repository from the development repository, and never directly from the build infrastructure.
+This step also minimizes the risk of faulty software affecting end users.
 This repository holds all of the software environment which is visible to the end users on all of our clusters.
-Once a piece of software is deployed onto this repository, it becomes visible on all clusters within 30 minutes.
+Once software is published to this repository, it becomes visible on all clusters within about 10 minutes.
 
 Both repositories above are available publicly, and can be mounted on any cluster or virtual machine in the cloud.
-We also have a third repository which is used to host software packages that are restricted to Compute Canada due to license limitations.
+There is a third repository, also hosted on the production servers, which provides software packages that are restricted to Compute Canada due to license limitations.
 We work with software vendors and developers in order to get the authorization to install licensed and commercial software packages in this repository.
 The most common licensing model that we use is a BYOL (bring your own license) model, in which we provide the software package, but end users provide a license server to connect to.
-Another model that is used is to control access to a software package via POSIX groups.
-By default, CVMFS clients do not honour POSIX group and filesystem permissions, but they can be configured to do so.
-Since this kind of restriction requires a specific LDAP and a specific configuration, this repository is restricted and can only be made available to clusters that are run by Compute Canada teams.
+Another model is to control access to a software package via POSIX groups.
+By default, the repository content on a CVMFS client belongs to a \texttt{cvmfs} service account and group, but this can be changed so that the original ownership metadata as published on the stratum 0 is used, thereby only allowing access to members of authorized POSIX groups.
+Since this relies on a specific LDAP domain and client configuration, access to this repository is restricted by network ACLs on the stratum 1 servers, and can only be granted to clusters that are run by Compute Canada teams.
 
-One final use case for CVMFS is to hold and distribute datasets.
-However, default CVMFS configuration is optimized for a large number of small files, rather than a small number of large files.
-This means that a special configuration must be done for this repository.
-At the time of writing this paper, this use case is still experimental within Compute Canada, but other groups are known to use CVMFS in this way.
+There is a fourth CVMFS repository that distributes configuration files instead of software.
+This special configuration repository resembles \texttt{/etc/cvmfs/} and provides information (such as URLs and public keys) required by CVMFS clients in order to access all {\it other} repositories.
+The benefit of this approach is that most CVMFS configuration can be centrally managed, and updates can be immediately pushed to all clients in order to set up a new repository or commission or decommission a stratum 1 server, rather than having to wait for all clients across the country to be updated.
+
+One final use case for CVMFS is to distribute datasets.
+By default, CVMFS  is optimized for a large number of small files, rather than a small number of large files, so different  configurations and access methods are needed for data repositories.
+At the time of writing this paper, this is used by some of our research groups in genomics, but not directly by our staff.
 
@@
 The second layer is what we call our compatibility layer.
-It contains almost all packages that one would typically install with the Linux distribution's package manager, such as {\it yum} or {\it apt-get}.
+It contains almost all packages that one would typically install with the Linux distribution's package manager, such as \texttt{yum} or \texttt{apt-get}.
 
@@
 The installation path of each software package installed through Nix resides in a path called the Nix store.
-This stores contains a single directory for each version of each software package installed through Nix.
+This store contains a single directory for each version of each software package installed through Nix.
 These directories are named after a hash calculated from the installation recipe and its dependencies.
@@
 In addition to the store, Nix creates directories that are called profiles.
-These profile directories contain a farm of symbolic links that point to files in the Nix store.
-For example, it will contain a {\it bin} directory, which will contain symbolic links to typical operating system commands such as {\it ls} or {\it gcc}, which will reside in different paths in the store.
-It will also contain {\it lib} and {\it include} directories which will contain symbolic links to libraries and header files located in the store.
+These profile directories contain a forest of symbolic links that point to files in the Nix store.
+For example, it contains a \texttt{bin/} directory, which contains symbolic links to typical operating system commands such as \texttt{ls} and \texttt{gcc}, which reside in different paths in the store.
+A profile directory also contains \texttt{lib/} and \texttt{include/} directories which contain symbolic links to libraries and header files located in the store.
 
-Profiles are the paths that are exposed to the end users, which is done by adding paths to their {\it PATH}, {\it CPATH} and other similar environment variables.
+Profiles are the paths that are exposed to the end users, which is done by adding paths to their \texttt{PATH}, \texttt{CPATH} and other similar environment variables.
 Each profile ever created is kept, but users are only exposed to the most recent one.
-Each profile represents the complete state of the user-exposed environment at a given time in the past.
+Each profile represents the complete state of the user-exposed environment at a given time.
 
@@
 
-This works well until a software exposes paths to the Nix store directly, instead of the paths to Nix profiles.
-We call this a Nix store {\it leak}.
+This works well until a software package exposes paths to the Nix store directly, instead of the paths to Nix profiles.
+We call this a Nix store leak.
 This happens for example with Python when using virtual environments.
-Virtual environments in Python copy the {\it python} executable to a secondary location, outside of Nix.
-However, all binaries installed by Nix use {\it RPATH} for shared object resolutions in the store directly.
+Virtual environments in Python copy the \texttt{python} executable to a secondary location, outside of Nix.
+However, all binaries installed by Nix use \texttt{RPATH} for shared object resolutions in the store directly.
 This means that copying a binary outside of Nix creates a dependency between a binary that now resides outside of the store, and the store itself.
 Because Nix is unaware of that dependency, any update of one of the dependencies followed by garbage collection or selective deployment of files will end up breaking this binary.
-We have encountered this issue with Python, Perl and Qt, which we have therefore moved outside of Nix and installed them through EasyBuild instead.
+We have encountered this issue with Python, Perl and Qt, which we have therefore moved outside of Nix and installed through EasyBuild instead.
 
-Another issue that we have encountered is that many Nix recipes call for installing sub-packages of a package in separate directories.
-For example, for Qt, {\it qmake} and {\it OTHER COMMAND} reside in different directories in the Nix store.
-This sometimes cause issues when installing a dependent package which assumes that its dependencies are installed in a more traditional fashion, with all of the binaries in the same {\it bin} folder, alongside with headers in the {\it include} folder, and libraries in the {\it lib} folder.
-This has mostly been an issue for Qt.
+Also, we have found that many Nix recipes call for installing sub-packages of a package in separate directories.
+For example, for Qt, \texttt{qmake} and \texttt{qtpaths} or \texttt{qtdiag} reside in different directories in the Nix store.
+This sometimes causes issues when installing a dependent package which assumes that its dependencies are installed in a more traditional fashion, with all of the binaries in the same \texttt{bin/} directory, alongside headers in the \texttt{include/} directory and libraries in the \texttt{lib/} directory.
 
@@
 Gentoo Prefix~\cite{Gentoo} is an alternative solution to provide a compatibility layer.
-It allows installation of most packages that form a typical Linux operating system in a non-default location.
-Contrary to Nix however, packages are installed in a standard directory structure ({\it bin}, {\it lib}, {\it include}, {\it share}).
+It allows installation of most packages that form a typical Linux operating system, in a non-default location.
+Contrary to Nix however, packages are installed in a standard directory structure (\texttt{bin/}, \texttt{lib/}, \texttt{include/}, \texttt{share/}).
 This is helpful when updates are required, and minimizes problems with applications that expect a standard directory structure.
@@
 Scientific applications constitute the majority of our software stack.
-Since their performance is critical, they can not be installed in the compatibility layer, as they need to be optimized for specific CPU instructions.
+Since their performance is critical, they can not be installed in the compatibility layer, as they need to be optimized for specific CPU instruction sets.
 To install them, we use the EasyBuild~\cite{EasyBuild2012,EasyBuild2014,EasyBuild2016} software installation tool.
@@
 It is designed to create optimized binaries, and eases recompilation of multiple versions of a given application using many combinations of compilers or libraries.
-EasyBuild is composed of three parts: the framework, easyblocks, and easyconfig files (also known as easyconfigs).
+EasyBuild is composed of three parts: the framework, easyblocks, and easyconfig files (also known as ``easyconfigs'').
 
-The framework contains all of the core code which drives the logic of installing a software.
-It drives all of the steps necessary to install a software: parsing the configuration options, downloading source files, patching them, configuring, building and installing the software, as well as performing post-installation steps such as sanity checks, generating a module or running custom commands.
+The framework contains all of the core code which drives the logic of installing a software package.
+It includes the following steps: parsing the configuration options, downloading source files, patching them, configuring, building and installing the software, as well as performing post-installation steps such as sanity checks, generating a module or running custom commands.
 
 Easyblocks are Python classes that define the steps required to install a given software package.
-There are generic easyblocks, such as {\it ConfigureMake} or {\it CMakeMake}, as well as custom easyblocks for packages that have more complicated installation procedures, such as OpenFOAM or PETSc.
+There are generic easyblocks, such as \texttt{ConfigureMake} or \texttt{CMakeMake}, as well as custom easyblocks for packages that have more complicated installation procedures, such as OpenFOAM or PETSc.
 
-Easyconfig files parametrize the execution of the code from the easyblocks.
-They define the software name and version, the URL from which the source files can be downloaded as well as options to be used for the various steps.
+Easyconfig files parameterize the execution of the code from the easyblocks.
+They define the software name and version, the URL from which the source files can be downloaded, and options to be used for the various steps.
 
@@
 They represent a set or subset of a compiler, an MPI implementation, mathematical libraries and CUDA.
-Having the ability to generate multiple builds of a given software simply by providing a command line argument to change the toolchain has proven invaluable to our workflow, and has allowed us to recompile hundreds of packages in matter of days when deploying a new version of compiler or a cluster with a new CPU architecture.
+Having the ability to generate multiple builds of a given software simply by providing a command line argument to change the toolchain has proven invaluable to our workflow, and has allowed us to recompile hundreds of packages in a matter of days when deploying a new version of compiler or a cluster with a new CPU architecture.
 
@@
 This limits the list of visible modules which could otherwise be very extensive.
-This can however throw off some users, since they don’t automatically see all available modules.
+This can however confuse some users, since they don’t automatically see all available modules.
 Lmod~\cite{Lmod} alleviates this issue by providing a separate command to crawl the full module tree and instructs users on how to load the module they want.
@@
 For scripting languages that support installing packages, such as Python, R or Perl, we leverage the existing package manager of that language, rather than installing them as modules.
-In the case of Python, we are building Python wheels~\cite{Wheels}, linking against our software stack.
-Wheels are binary packages that can be installed in a matter of seconds, similar to CentOS RPM packages.
+In the case of Python, we build Python wheels~\cite{Wheels}, linked against our software stack.
+Wheels are binary packages that can be installed in a matter of seconds, similar to RPM packages.
 Since they are built by our team, and linked against our optimized software stack, we ensure that they are easy to install and offer good performance.
-
-\subsection{Licensed and restricted software packages}
-\label{sub:Licensed_software_packages}
-Compute Canada also has agreements with a number of software vendors or developers of applications that require a license.
-In these cases, we install the package using the same tools, but it is deployed on a separate CVMFS repository which is only accessible to Compute Canada clusters.
 
@@
 \label{sub:Documentation}
-Every software package that is installed through EasyBuild get automatically listed on our documentation website, at \\
+Every software package that is installed through EasyBuild gets automatically listed on our documentation website:
+\begin{center}
     \url{https://docs.computecanada.ca/wiki/Available_software}
-A similar list of available Python wheels is automatically generated at \\
+\end{center}
+Similarly, a list of available Python wheels is automatically generated:
+\begin{center}
 \url{https://docs.computecanada.ca/wiki/Available_wheels}    
+\end{center}
 
@@
 \label{sub:Logging}
-Our module system will automatically log (through {\it logger}) each module that is loaded.
-The information is logged in an easily parsable list of key-value pairs, which can then be aggregated into your favorite monitoring tool such as Elastic Search.
+Our module system will automatically log (through \texttt{logger}) each module that is loaded.
+The information is logged in an easily parsable list of key-value pairs, which can then be aggregated into your favorite monitoring tool, such as Elastic Search.
 
-\section{Making the software environment available on your cluster or computer}
+\section{Making the software environment available on your system}
 \label{sec:Making_it_available}
@@
 
-After CVMFS is configured, one can simply execute the command below in order to obtain Compute Canada's software environment \\
-\texttt{source /cvmfs/soft.computecanada.ca/config/profile/bash} \\
+After CVMFS is configured, one can simply execute this command in order to set up Compute Canada's software environment:
 
-By default, our default set of modules will try to detect the most recent set of instructions that is supported by the CPU amongst {\it SSE3}, {\it AVX}, {\it AVX2} and {\it AVX512}.
-It will also try to detect whether the hardware has a {\it Mellanox InfiniBand}, {\it Intel OmniPath}, or an Ethernet network fabric.
-If these are not detected correctly, or if you have a cluster made up of mixed hardware, you can overwrite these defaults by defining the {\verb!RSNT_ARCH!} or {\verb!RSNT_INTERCONNECT!} environment variables before executing the command above.
+\begin{center}
+\texttt{source \textbackslash \linebreak
+/cvmfs/soft.computecanada.ca/config/profile/bash.sh}
+\end{center}
 
-Note that the first time a command is ran, it will take some longer time to start the command, as the CVMFS client is fetching the files that are needed to run the command from our stratum 1 servers.
-Following executions of the same command will however be fast as files will have been cached locally.
+The default set of modules will automatically try to detect the most recent set of instructions that is supported by the CPU amongst  SSE3, AVX, AVX2 and AVX512.
+It will also try to detect whether the hardware has a Mellanox InfiniBand, Intel OmniPath, or Ethernet network fabric.
+If these are not detected correctly, or if you have a cluster made up of mixed hardware, you can overwrite these defaults by defining the \texttt{RSNT\_ARCH} or \texttt{RSNT\_INTERCONNECT} environment variables before executing the command above.
 
-Also note that the command above will \textbf{not} import our environment for any user ID lower than 1000, which are typically privileged users on a Linux system.
-In order to force the command above to execute nevertheless, you can create a file named \verb!.force_cc_cvmfs!
-in the user's \verb!$HOME!
-directory.
+Note that the first time any software or data is accessed, it will take longer to load, as the CVMFS client is fetching the content from our stratum 1 servers.
+However, subsequent access will be much faster once the content is cached locally.
+Also note that the command above will {\it not} import our environment for any user IDs lower than 1000, which are typically privileged users on a Linux system.
+In order to force the command above to execute nevertheless, you can create a file named \texttt{.force\_cc\_cvmfs} in the user's \texttt{\$HOME} directory.
+Conversely, a \texttt{\$HOME/.skip\_cc\_cvmfs} file can be created to exclude a user from setting up the software environment.
+This is recommended for system administrators so their environment does not depend on CVMFS, especially if they are doing operations which may interrupt access to CVMFS, such as updating or reconfiguring it.
 
-Making our software environment available on a cluster requires the additional step of configuring one or multiple HTTP proxies.
-This is required to avoid overloading the stratum-1 server by having a large number of nodes all fetching the same files at the same time.
-If you are interested in testing or making our software environment available to your users on your cluster, we ask that you first contact us.
+If you plan to use our CVMFS repositories on a large number of nodes such as a cluster, we require that you deploy one or more caching HTTP proxies, to avoid overloading the stratum 1 servers.
+If you are interested in testing or making our software environment available to your users on your cluster, we ask that you first contact us by writing to
+\begin{center}
+\url{rsnt-software@computecanada.ca}    
+\end{center}
 Note that this software stack comes with absolutely no warranty.
@@
 \label{sec:Conclusion}
-We have built a software environment that is portable, is optimized for multiple CPU architectures and network interconnect, and can be deployed in a matter of minutes on a single node, and barely longer for a whole cluster.
+We have built a software environment that is portable, is optimized for multiple CPU architectures and network interconnects, and can be deployed in a matter of minutes on a single node, and not much longer for a whole cluster.
 The design of the infrastructure guarantees that the software environment seen by our end users is the same across all of our clusters, and will not diverge over time.
@@
 It can also help support infrastructures owned by various institutions.
+This is the case of many universities in Canada that used to host a cluster which is now used by the institution only, or universities that deploy new local infrastructure and don't have the human resources to build a large software stack.
 
@@
 
-%
 % The next two lines define the bibliography style to be used, and the bibliography file.

### 2019-04-08 paper.tex
Reworked introduction.

@@
 The second principle is that the software environment should be decoupled from the underlying operating system.
+Since each tool for software package installation has strengths and weaknesses, we adopted a layered approach to benefit from the best tool for each kind of software package.
+We briefly described this in Section~\ref{sec:Layered_environment}.
+
 When we first designed the solution, new clusters were being deployed with the CentOS 7 operating system.
@@
 For scientific packages, we also wanted to ensure that the software packages were optimized for specific hardware architectures, which Nix cannot do easily.
-The tool we chose for this task is EasyBuild~\cite{EasyBuild2012,EasyBuild2014,EasyBuild2016}, which we will discuss in Section~\ref{sub:Compiled_software_packages}.
+The tool we chose for this task is EasyBuild~\cite{EasyBuild2012,EasyBuild2014,EasyBuild2016}, which we will discuss in Section~\ref{sec:Scientific_applications}.
 
@@
 Using a hierarchical module tree, which is possible in a newer implementation of environment modules called Lmod \cite{Lmod}, helps address this problem.
-We will discuss how our implementation meets the third and fourth principles in Section~\ref{sec:Scientific_applications}.
+We will discuss how our implementation meets the third and fourth principles in Section~\ref{sub:Compiled_software_packages}.
+
+Finally, we describe in Section~\ref{sec:Making_it_available} how you can make use of our software stack on your computer or cluster, mention other marginal benefits from our design in Section~\ref{sec:Other_benefits}.
 

### 2019-04-09 paper.tex
Merged in new changes from Ryan and Maxime

@@
 The second principle is that the software environment should be decoupled from the underlying operating system.
-Since each tool for software package installation has strengths and weaknesses, we adopted a layered approach to benefit from the best tool for each kind of software package.
-We briefly described this in Section~\ref{sec:Layered_environment}.
+Since each tool for software package installation has strengths and weaknesses, we adopted a layered approach to leverage the best tool for each kind of software package.
+We briefly describe this in Section~\ref{sec:Layered_environment}.
 
@@
 
-Finally, we describe in Section~\ref{sec:Making_it_available} how you can make use of our software stack on your computer or cluster, mention other marginal benefits from our design in Section~\ref{sec:Other_benefits}.
+Finally, we describe in Section~\ref{sec:Making_it_available} how you can use our software stack on your computer or cluster, and summarize the benefits of our design in Section~\ref{sec:Conclusion}.
 
@@
 
-One final use case for CVMFS is to distribute datasets.
+One additional use case for CVMFS is to distribute datasets.
 By default, CVMFS  is optimized for a large number of small files, rather than a small number of large files, so different  configurations and access methods are needed for data repositories.
@@
 
-The framework contains all of the core code which drives the logic of installing a software package.
-It includes the following steps: parsing the configuration options, downloading source files, patching them, configuring, building and installing the software, as well as performing post-installation steps such as sanity checks, generating a module or running custom commands.
+The framework contains all of the core code which drives the logic of the entire software installation process, including parsing the configuration options, downloading source files, patching them, configuring, building and installing the software, as well as performing post-installation steps such as sanity checks, generating a module or running custom commands.
 
@@
 We use a version of a hierarchical naming scheme, in which modules are organized in a tree structure, hiding those that are incompatible with currently loaded modules.
-This limits the list of visible modules which could otherwise be very extensive.
+This limits the list of visible modules, which could otherwise be very extensive.
 This can however confuse some users, since they don’t automatically see all available modules.
@@
 \label{sec:Making_it_available}
-Making our software environment available on a single computer is as simple as installing the CVMFS client and configuring repositories manually or installing our RPM
-\textbf{TODO: Document our RPM}
+Making our software environment available on a single computer is as simple as installing the CVMFS client and configuring repositories manually or installing our RPM.
+This is described on this webpage:
+\begin{center}
+    \url{https://docs.computecanada.ca/wiki/Accessing_our_software_environment_on_your_computer}
+\end{center}
 
@@
 If you plan to use our CVMFS repositories on a large number of nodes such as a cluster, we require that you deploy one or more caching HTTP proxies, to avoid overloading the stratum 1 servers.
-If you are interested in testing or making our software environment available to your users on your cluster, we ask that you first contact us by writing to
-\begin{center}
-\url{rsnt-software@computecanada.ca}    
-\end{center}
+If you are interested in testing or making our software environment available to your users on your cluster, we ask that you first contact us as described in the webpage mentioned before.
 Note that this software stack comes with absolutely no warranty.
 
-\section{Other benefits and conclusion}
+\section{Conclusion}
 \label{sec:Conclusion}
@@
 The design of the infrastructure guarantees that the software environment seen by our end users is the same across all of our clusters, and will not diverge over time.
-This makes for a better user experience, as it becomes much easier to migrate between two clusters.
+This makes it much easier for users to use multiple clusters or migrate between clusters, improving the user experience.
 It also reduces the workload on our staff by unifying the installation process for all clusters, and makes supporting our users easier because all clusters work the same way.
 
-Because of the portability of our software stack and the distribution mechanism, it can be used on any cluster, or any virtual machine on a cloud.
+Because of the portability of our software stack and the distribution mechanism, it can be used on any cluster, or any Linux computer or virtual machine on a cloud.
 For example, for training events, we have been using it to provision small replicas of Compute Canada clusters in short-lived virtual clusters deployed on our OpenStack platform.
 We have also used this portable software environment to expose a cluster environment directly through a browser, via JupyterHub.
-It can also help support infrastructures owned by various institutions.
-This is the case of many universities in Canada that used to host a cluster which is now used by the institution only, or universities that deploy new local infrastructure and don't have the human resources to build a large software stack.
+
+Since software installation is no longer tightly coupled to any particular site or system, this approach has reduced the time required to commission major new Compute Canada clusters and make them accessible to our users by several months.
+At the same time, it also benefits universities and other research organizations running smaller or older clusters, desktops, and laboratory computing infrastructure, which may not have the human resources to build and maintain a comprehensive research software stack.
 

### 2019-04-09 paper.tex
Merged in Bart's changes.

@@
 \usepackage[utf8]{inputenc}
+\usepackage{colortbl}
+\usepackage{amssymb}
 %
@@
 % The "title" command has an optional parameter, allowing the author to define a "short title" to be used in page headers.
-\title{Providing a Unified Software Environment for Canada’s National Advanced Computing Centers}
+\title[A Unified Software Environment for Canada’s Advanced Computing Centers]{Providing a Unified Software Environment for Canada’s National Advanced Computing Centers}
 
@@
 \begin{figure*}
-  \includegraphics[width=0.7\textwidth]{layers.png}
+  \begin{center}
+  {\Large
+  \begin{tabular}{|p{0.7\textwidth}|}
+  \hline \cellcolor{cyan}
+  EasyBuild layer: modules for Intel, PGI, Open MPI, MKL, high-level applications.
+  Multiple instruction sets (SSE3, AVX, AVX2, AVX512).\newline
+  \texttt{/cvmfs/soft.computecanada.ca/easybuild/\{modules,software\}/2017}\\
+  \hline \cellcolor{green}
+  EasyBuild-generated modules around Nix profiles: for example Clang, Eclipse,
+  Erlang, FPC, GCC, Ruby.\newline
+  \texttt{/cvmfs/soft.computecanada.ca/nix/var/nix/profiles/[a-z]\**}\\
+  \hline \cellcolor{yellow}
+  Nix layer: GNU C Library (glibc), autotools, make, bash, cat, ls, awk, grep, etc. \newline
+  \texttt{module load nixpkgs/16.09 $\Rightarrow$ \$EBROOTNIXPKGS=} \newline
+  \texttt{/cvmfs/soft.computecanada.ca/nix/var/nix/profiles/16.09}\\
+  \hline \cellcolor{lightgray} 
+  Gray area: Lustre client libraries, IB/OmniPath/InfiniPath client libraries 
+  (all dependencies of Open MPI). In Nix layer, but can be overridden using
+  \texttt{PATH} and \texttt{LD\_LIBRARY\_PATH}.\\
+  \hline \cellcolor{pink}
+  OS kernel, daemons, drivers, libcuda, anything privileged (e.g. the sudo
+  command): always local. Some legally restricted software too, such as VASP.\\
+  \hline
+  \end{tabular}
+  }
+  \end{center}
   \caption{Layered structure of our software stack.
-The bottom layer (red) is entirely managed by the system administrators of each cluster.
+The bottom layer (pink) is entirely managed by the system administrators of each cluster.
 The second layer (grey) contains libraries that can be installed centrally, through Nix, but can be overridden locally.
@@
 They are installed through Nix, but wrapped into a module by EasyBuild.
-The top layer (blue) contains packages that are not typically installed in a Linux distribution, and are installed via EasyBuild.}
+The top layer (cyan) contains packages that are not typically installed in a Linux distribution, and are installed via EasyBuild.}
   \label{tab:layers}
@@
 This sometimes causes issues when installing a dependent package which assumes that its dependencies are installed in a more traditional fashion, with all of the binaries in the same \texttt{bin/} directory, alongside headers in the \texttt{include/} directory and libraries in the \texttt{lib/} directory.
+
+In general, Nix provides an environment that is internally consistent but is best developed using Nix itself, through tools such as \texttt{nix-shell}.
+Such tools can only be used if the Nix store is writable, an impossibility with CVMFS.
+Attempting to use Nix in conjunction with other development tools using environment variables can be fragile at times.
+Moreover, the dependency of the hash on the whole installation recipe, instead of only the dependencies (as, for example, implemented in Spack~\cite{Spack}), causes the store path to change even if a bug fix or security update needs to be applied, aggravating this problem.
+We must note however, that some clusters in the Compute Canada provide Nix with a writable Nix store, which is completely separate from the software stack we are describing here.
 
@@
 If you plan to use our CVMFS repositories on a large number of nodes such as a cluster, we require that you deploy one or more caching HTTP proxies, to avoid overloading the stratum 1 servers.
-If you are interested in testing or making our software environment available to your users on your cluster, we ask that you first contact us as described in the webpage mentioned before.
+If you are interested in testing or making our software environment available to your users on your cluster, we ask that you first contact us as described in the web page mentioned before.
 Note that this software stack comes with absolutely no warranty.

### 2019-04-10 paper.tex
Changed wiki URL
Merged Ryan's comments.

@@
 The second principle is that the software environment should be decoupled from the underlying operating system.
-Since each tool for software package installation has strengths and weaknesses, we adopted a layered approach to leverage the best tool for each kind of software package.
+Since each tool for software package installation has different strengths and weaknesses, we adopted a layered approach to leverage the best tool in each situation.
 We briefly describe this in Section~\ref{sec:Layered_environment}.
@@
 CVMFS is designed similarly to the Network Time Protocol stratum model, as illustrated in Figure~\ref{fig:CVMFS_structure}.
-There is one source of truth, the stratum 0 server, where software is injected.
+There is one source of truth, the stratum 0 server, where software is published.
 The stratum 0 is never accessed directly by clients.
-If it fails, the only immediate consequence is that injection of new files is not possible until it is back online.
+If it fails, the only immediate consequence is that publishing new files is not possible until it is back online.
 
@@
 Internally, CVMFS uses a content-addressable storage scheme, wherein files exceeding a certain size are broken into smaller objects, and each object is hashed, compressed, and stored in a location defined by its hash, resulting in automatic deduplication.
-Hashes also allow CVMFS clients to quickly verify the integrity of content received over the network, and invalidate their local caches when content is changed, by means of a Merkle tree.
+Hashes also allow CVMFS clients to quickly verify the integrity of content received over the network, and quickly switch their repositories to new revisions when updates are published,  by means of a Merkle tree.
 
@@
 This step also minimizes the risk of faulty software affecting end users.
-This repository holds all of the software environment which is visible to the end users on all of our clusters.
+This repository holds the software environment which is visible to the end users on all of our clusters.
 Once software is published to this repository, it becomes visible on all clusters within about 10 minutes.
@@
 By default, CVMFS  is optimized for a large number of small files, rather than a small number of large files, so different configurations and access methods are needed for data repositories.
-At the time of writing this paper, this is used by some of our research groups in genomics, but not directly by our staff.
+At the time of writing this paper, data repositories are used by some of our research groups in genomics, but not directly by our staff yet.
 
@@
 
-The second layer is what we call our compatibility layer.
-It contains almost all packages that one would typically install with the Linux distribution's package manager, such as \texttt{yum} or \texttt{apt-get}.
+The second and third layers constitute what we call our compatibility layer, containing almost all packages that one would typically install with the Linux distribution's package manager, such as \texttt{yum} or \texttt{apt-get}.
 
-The third layer is the scientific application layer.
-It contains mostly scientific or performance critical applications, including for example MPI implementations, the CUDA toolkit, mathematical libraries, and domain specific applications.
+The fourth and fifth layers are the scientific application layers.
+They contain mostly scientific or performance critical applications, including for example MPI implementations, the CUDA toolkit, mathematical libraries, and domain specific applications.
 
-\begin{figure*}
+\begin{table*}
   \begin{center}
@@
   \caption{Layered structure of our software stack.
-The bottom layer (pink) is entirely managed by the system administrators of each cluster.
-The second layer (grey) contains libraries that can be installed centrally, through Nix, but can be overridden locally.
+The bottom layer (pink) is entirely managed by the system administrators of each cluster and is not on CVMFS.
+The second layer (grey) contains libraries that are available via Nix on CVMFS, but can be overridden locally.
 The third layer (yellow) contains everything that one would typically find in the Linux user space.
@@
   \label{tab:layers}
-\end{figure*}
+\end{table*}
 
@@
 We exclude the Linux kernel and its modules, hardware drivers, and privileged binaries from this layer.
+However, some libraries such as InfiniBand client libraries that are close to the hardware are also available in this layer (the gray area in Table~\ref{tab:layers}) and can optionally be used, or can be provided locally outside CVMFS.
 
@@
 These directories are named after a hash calculated from the installation recipe and its dependencies.
-This creates a very large number of subdirectories, many of which contain almost identical files.
+This creates a very large number of subdirectories, many of which contain nearly identical files.
 
@@
 These profile directories contain a forest of symbolic links that point to files in the Nix store.
-For example, it contains a \texttt{bin/} directory, which contains symbolic links to typical operating system commands such as \texttt{ls} and \texttt{gcc}, which reside in different paths in the store.
+For example, each profile contains a \texttt{bin/} directory, which contains symbolic links to typical operating system commands such as \texttt{ls} and \texttt{gcc}, which reside in different paths in the store.
 A profile directory also contains \texttt{lib/} and \texttt{include/} directories which contain symbolic links to libraries and header files located in the store.
+We provide one main profile as the compatibility layer and some custom profiles for applications from that layer for which we provide multiple versions via modules that are generated by EasyBuild.
+Those custom profiles hence provide a hybrid between the top layer and the compatibility layer.
 
@@
 Each profile ever created is kept, but users are only exposed to the most recent one.
-Each profile represents the complete state of the user-exposed environment at a given time.
+Each profile represents the complete state of the user-accessible environment at a given time.
 
@@
 Virtual environments in Python copy the \texttt{python} executable to a secondary location, outside of Nix.
-However, all binaries installed by Nix use \texttt{RPATH} for shared object resolutions in the store directly.
+However, all binaries installed by Nix use \texttt{RPATH} for shared object resolution directly in the store.
 This means that copying a binary outside of Nix creates a dependency between a binary that now resides outside of the store, and the store itself.
@@
 Moreover, the dependency of the hash on the whole installation recipe, instead of only the dependencies (as, for example, implemented in Spack~\cite{Spack}), causes the store path to change even if a bug fix or security update needs to be applied, aggravating this problem.
-We must note however, that some clusters in the Compute Canada provide Nix with a writable Nix store, which is completely separate from the software stack we are describing here.
+We must note however, that some clusters in Compute Canada provide Nix with a writable Nix store, which is completely separate from the software stack described here.
 
@@
 
-The framework contains all of the core code which drives the logic of the entire software installation process, including parsing the configuration options, downloading source files, patching them, configuring, building and installing the software, as well as performing post-installation steps such as sanity checks, generating a module or running custom commands.
+The framework contains the core code which drives the logic of the entire software installation process, including parsing the configuration options, downloading source files, patching them, configuring, building and installing the software, as well as performing post-installation steps such as sanity checks, generating a module or running custom commands.
 
@@
 \label{sec:Making_it_available}
-Making our software environment available on a single computer is as simple as installing the CVMFS client and configuring repositories manually or installing our RPM.
-This is described on this webpage:
+To access the Compute Canada repositories on your own computer, you can install and configure the CVMFS client as described here:
 \begin{center}
-    \url{https://docs.computecanada.ca/wiki/Accessing_our_software_environment_on_your_computer}
+    \url{https://docs.computecanada.ca/wiki/Accessing_CVMFS}
 \end{center}
 
-After CVMFS is configured, one can simply execute this command in order to set up Compute Canada's software environment:
+Then, to set up the Compute Canada software environment, simply execute this command:
 
@@
 \end{center}
+
+If you plan to use our CVMFS repositories on a large number of nodes such as a cluster, or to present our software environment to your users, we ask that you first contact us as described in the documentation web page above.
+Also, if you set up a large number of client nodes we require that you deploy them behind one or more caching HTTP proxies, to avoid overloading the stratum 1 servers.
+Note that this software stack is provided as is and comes with absolutely no warranty.
 
@@
 This is recommended for system administrators so their environment does not depend on CVMFS, especially if they are doing operations which may interrupt access to CVMFS, such as updating or reconfiguring it.
-
-If you plan to use our CVMFS repositories on a large number of nodes such as a cluster, we require that you deploy one or more caching HTTP proxies, to avoid overloading the stratum 1 servers.
-If you are interested in testing or making our software environment available to your users on your cluster, we ask that you first contact us as described in the web page mentioned before.
-Note that this software stack comes with absolutely no warranty.
 

### 2019-05-21 paper.tex
Added AMC provided information

@@
 \setcopyright{acmlicensed}
-\acmConference[PEARC19]{PEARC19: Practice and Experience in Advanced Research Computing}{July 28 -- August 01, 2019}{Chicago, IL}
-%\acmBooktitle{Woodstock '18: ACM Symposium on Neural Gaze Detection, June 03--05, 2018, Woodstock, NY}
-%\acmPrice{15.00}
-%\acmDOI{10.1145/1122445.1122456}
-%\acmISBN{978-1-4503-9999-9/18/06}
+\acmConference[PEARC '19]{Practice and Experience in Advanced Research Computing}{July 28-August 1, 2019}{Chicago, IL, USA}
+\acmBooktitle{Practice and Experience in Advanced Research Computing (PEARC '19), July 28-August 1, 2019, Chicago, IL, USA}
+\acmPrice{15.00}
+\acmDOI{10.1145/3332186.3332210}
+\acmISBN{978-1-4503-7227-5/19/07}
+
+%\CopyrightYear{2019}
+%\setcopyright{acmlicensed}
+%\conferenceinfo{PEARC '19,}{July 28-August 1, 2019, Chicago, IL, USA}
+%\isbn{978-1-4503-7227-5/19/07}\acmPrice{\$15.00}
+%\doi{https://doi.org/10.1145/3332186.3332210}
 
